---
title: "Report on life expectancy"
author: "Corinne Riddell"
date: "August 28, 2019"
output:
  pdf_document: default
  html_document: default
---

```{r lib-load, echo = FALSE, message=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(glue)
```

```{r choose-your-state, echo=FALSE}
my_state <- "New York"
```

# Life expectancy for non-Hispanic Black and White men and women in `r my_state` between 1969-2013

### Introduction 

Life expectancy is one of the core measures used in public health to comment on 
the well-being of groups of people. Differences in life expectancy by race/ethnicity,
for individuals living in the same region can reflect underlying inequalities in
policies, access to care, food environments, structural and systemic racism, among
other potential causes.

### Research objective (Problem)

The purpose of this short report is to visualize life expectancy among Black and 
White men and women in `r my_state` between 1969 and 2013. 

### Plan

Death certificates in the United States include race/ethnicity, age at death, 
and date of death and capture all deaths of US residents. These data are aggregated by the 
[CDC's National Cancer Institute](https://seer.cancer.gov/seerstat/) into the 
SEER*Stat software. Previously, Riddell et al.^1^, analyzed these data to 
compute estimated trends in life expectancy for non-Hispanic Black and White men 
and women, for 40 US states between 1969 and 2013. States without enough data were
excluded from these analyses.

To carry out this short report, we will use data from Riddell et al. to 
visualize trends in life expectancy as part of an exploratory data analysis. In 
particular, we will plot time trends for Black and White men and women in 
`r my_state`.

### Data

Here are the first five rows of these data for `r my_state`:
```{r import-and-filter-data, echo=F, message=F}
le_data <- read_csv("../Data/Life-expectancy-by-state-long.csv")

my_state_data <- le_data %>% 
  filter(state == my_state) %>%
  mutate(gender = case_when(sex == "Female" ~ "Women",
                            sex == "Male" ~ "Men")) %>%
  select(-sex)

head(my_state_data)
```

### Analysis

```{r first-look-time-plots, echo=F, fig.width=12}
ggplot(data = my_state_data, aes(x = year, y = LE)) + 
  geom_line(aes(col = race)) + 
  facet_wrap(~ gender) +
  labs(y = "Life expectancy (years)",
       title = glue("Trends in life expectancy for Black and White men and women in {my_state}")
  ) +
  theme_minimal(base_size = 15)
```

```{r reshape-and-compute-LE-difference, echo=F, fig.width=12}
my_state_data_wide <- my_state_data %>% 
  spread(key = race, value = LE) %>%
  mutate(le_difference = white - black)

ggplot(my_state_data_wide, aes(x = year, y = le_difference)) + 
  geom_line() +
  facet_wrap(~ gender) +
  labs(y = "Difference in life expectancy (years)",
       title = glue("Difference in life expectancy between black and white men and women in {my_state}")
       ) +
  theme_minimal(base_size = 15)
```

### Conclusion

The difference in life expectancy in 1969 between non-Hispanic blacks and whites was `r my_state_data_wide %>% filter(year == 1969, gender == "Men") %>% select(le_difference) %>% round(1)` years for men and `r my_state_data_wide %>% filter(year == 1969, gender == "Women") %>% select(le_difference) %>% round(1)` for women in `r my_state`. 

By 2013, the difference was `r my_state_data_wide %>% filter(year == 2013, gender == "Men") %>% select(le_difference) %>% round(1)` years for men and `r my_state_data_wide %>% filter(year == 2013, gender == "Women") %>% select(le_difference) %>% round(1)` for women in `r my_state`.

### References

1. Riddell CA, Morrison KT, Harper S, Kaufman JS. Trends in the contribution of major causes of death to the black-white life expectancy gap by US state. **Health & Place**. 2018. 52:85-100. doi: 10.1016/j.healthplace.2018.04.003.  