---
title: "Bootstrap Confidence Intervals"
subtitle: (This material is not from the textbook)
author: "Corinne Riddell"
date: "November 6, 2019"
output: pdf_document
---

### Confidence interval recap

- So far we've (mostly) been using formulas of the form $\text{estimate}\pm\text{margin of error}$
to generate confidence intervals.

- We've applied these formulas to the mean $\bar{x}$ and the proportion $\hat{p}$. 
These "large sample" procedures use the Central Limit Theorem to calculate the 
standard error of the sampling distribution for $\bar{x}$ and $\hat{p}$

### Confidence interval recap

- We did this when the data met certain assumptions, one of which being that the
data were drawn from a population that is Normally distributed.

- Though there was some leeway around this assumption, what would happen if you 
were sure the assumption was violated? How would you calculate a 95% confidence
interval in that case?

- **We need a procedure to calculate confidence intervals that do not rely on the 
assumption of Normality.**

### Confidence intervals for other parameters

- What if we wanted a confidence interval for the median? Or a confidence interval
for the first quartile, $Q_1$, or some other parameter? 

- We couldn't use the same procedure because the CLT does not apply to the 
sampling distribution of the median or $Q1$, or to many other statistics.

- **We need a procedure to calculate confidence intervals for other parameters, 
such as the median or any other quartile or percentile.**

### Enter: The Bootstrap Confidence Interval

- The Bootstrap CI does not require the underlying distribution to be Normally 
distributed.

- The Bootstrap CI can be made for any parameter (based on its sample statistic).

- It was so-named because it uses only the information from the sample you have
to estimate the CI, such that the sample is "pulling itself up by its bootstraps".

- This method gained popularity in 1979 when popularized by Bradley Efron, and 
harnesses current computing power. 

- **It is a very popular method used to compute confidence intervals today** 

### Bootstrap example

This example came from bootstrap's advocate, Bradley Efron. Suppose we have 
measures of serum from 54 patients:

```{r, echo=FALSE, warning=F, message=F}
library(ggplot2)
library(dplyr)
```

```{r serum_data_input}
serum_data <- data.frame(serum = c(0.1, 0.1, 0.2, 0.4, 0.4, 0.6, 0.8, 0.8, 
                                   0.9, 0.9, 1.3, 1.3, 1.4, 1.5, 1.6, 1.6, 1.7,
                                   1.7, 1.7, 1.8, 2.0, 2.0, 2.2, 2.2, 2.2, 2.3, 
                                   2.3, 2.4, 2.4, 2.4, 2.4, 2.4, 2.4, 2.5, 2.5, 
                                   2.5, 2.7, 2.7, 2.8, 2.9, 2.9, 2.9, 3.0, 3.1, 
                                   3.1, 3.2, 3.2, 3.3, 3.3, 3.5, 4.4, 4.5, 6.4, 
                                   9.4))
```

```{r out.width="80%", fig.align='center', echo=F}
ggplot(serum_data, aes(serum)) + 
  geom_histogram(binwidth = 0.5, col = "white") + 
  theme_minimal(base_size = 15) +
  labs(title = "Histogram of serum measurements")
```

### First, calculate the theory-based 95% CI

Using the skills we already know, we can calculate the **95% CI for the mean** 
for these data using `dplyr`

```{r}
t_star <- qt(0.975, df = 53) # find t* to use to calculate the 95% CI

serum_data %>% summarise(mean_serum = mean(serum),
                         se_serum = sd(serum)/sqrt(n()),
                         lower_CI = mean_serum - t_star * se_serum,
                         upper_CI = mean_serum + t_star * se_serum)
```

Or by using the `t.test()` function:

```{r}
t.test(serum_data %>% pull(serum))
```

The 95% CI provides our best guess of where the true proportion lies. The  95% 
CI for $\mu$ is 1.90 to 2.74. We found this interval using a method that gives 
an interval that captures $\mu$ 19 times out of 20. 

### 95% CI for the median

- We can't use the above method to compute a 95% CI for the median because the 
CLT does not apply to the median. 
- However, we can use a method that relies on the assumption that the sample is 
a SRS from the underlying population

### How the Bootstrap CI is made

- If we truly have a SRS from the underlying population, this means that the 
distribution of serum in the sample should *approximate* the distribution of serum
in the population 
- Graphically, this means the shape of the histogram for the sample data should
approximate the shape of the density plot for the entire population 
- **The key: If we take repeated samples (with replacement) from our sample, we 
can approximate the sampling distribution for any statistic we'd like**
- This is the process of bootstrapping
- Let's apply this method to calculate the 95% CI for the median using the serum
data

### Boostrap confidence interval for the median

1. Calculate the median for the original sample of size 54. Denote this value by
$m$. This is our estimate of the median for the underlying population. We need to 
create a 95% CI around $m$.

```{r}
median_serum <- serum_data %>% summarise(median_serum = median(serum))
median_serum
```

2. Resample **with replacement** from the original sample a new sample, also of size 54.
3. Calculate the median based on resample #1. Call this median $m^*_1$.
4. Resample again. Calculate the median based on resample #2. Call this median $m^*_2$.
... repeat this resampling procedure several thousand times. 
5. Make a histogram of $m^*_1$, $m^*_2$, ... , $m^*_{1000}$. **This histogram 
approximates the sampling distribution for the median**
6. Calculate the bounds such that the middle 95% of the observations are between the
lower and upper bounds. In R, we can do this using `quantile(sample_median, 0.025)` and
`quantile(sample_median, 0.975)` to locate the 2.5th and 97.5th percentiles of the 
variable `sample_median`.

### Boostrap confidence interval for the median

This code resamples the data 1000 times and calculates the median for each
resample. It stores the median in a data frame called `many_sample_medians`

```{r}
# students, you don't need to know how this code works. 

number_of_bootstraps <- 10000
sample_size <- 54

calc_sample_stats <- function(df) {
  df %>% summarize(sample_median = median(serum))
}

set.seed(1)

many_sample_medians <- replicate(number_of_bootstraps, 
                                 sample_n(serum_data, sample_size, replace = T), 
                                 simplify = F) %>%
  lapply(., calc_sample_stats) %>%
  bind_rows() %>%
  mutate(sample.id = 1:n())

head(many_sample_medians)
```

### Calculate the lower and upper bounds of the 95% bootstrap CI

```{r}
# Understand this code. It takes the data frame containing the bootstrap
# sample medians and calculates the lower and upper CI using the quantile function.

bounds <- many_sample_medians %>%
  summarise(lower_CI = quantile(sample_median, 0.025), 
            upper_CI = quantile(sample_median, 0.975))

bounds
# how would the code change if we were interested in the 99% CI?
```

Thus, our best estimate of the median is 2.35. The bootstrapped 95% confidence 
interval for the median is 1.9 to 2.5.

### Plot the histogram of the bootstrapped medians and denote the 95% confidence interval

```{r, fig.align="centre", out.width="80%"}
ggplot(many_sample_medians, aes(x = sample_median)) + 
  geom_histogram(binwidth = 0.07, col = "white", fill = "#99d8c9") +
  labs(x = "Bootstrapped medians") + 
  geom_vline(aes(xintercept = median_serum %>% pull())) +
  theme_minimal(base_size = 15) +
  geom_vline(aes(xintercept = bounds %>% pull(lower_CI)), lty = 2) +
  geom_vline(aes(xintercept = bounds %>% pull(upper_CI)), lty = 2)
```

- Note that the sampling distribution for the median is not symmetric. It is skewed
left (for these data).
- The CI is not symmetric around our best guess for the sample mean (2.35). This is okay,
and even preferable.

### Another example: Calculate the bootstrap CI for the correlation coefficient

You have data on student demographics and scores across multiple subjects. You
are interested in the correlation between the scores for writing (`write`) and
mathematics (`math`)

```{r, out.width="80%", fig.align='center'}
hsb2 <- read.table("../Data/hsb2.csv", sep=",", header=T)

ggplot(data = hsb2, aes(write, math)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = F) +
  geom_text(aes(x = 40, y = 70, label = "corr = 0.62"), check_overlap = T) +
  theme_minimal(base_size = 15)
```

### Calculate the bootstrapped CI for the correlation

1. Calculate the correlation for the original sample of size 200. Denote this value by
$m$. This is our estimate of the median for the underlying population. We need to 
create a 95% CI around $m$.

```{r}
corr_math_write <- hsb2 %>% summarise(correlation = cor(write, math))
corr_math_write
```

2. Resample with replacement from the original sample a new sample, also of size 200.
3. Calculate the correlation based on resample #1. Call this correlation $r^*_1$.
4. Resample again. Calculate the correlation based on resample #2. Call this median $r^*_2$.
... repeat this resampling procedure several thousand times. 
5. Make a histogram of $r^*_1$, $r^*_2$, ... , $r^*_{10000}$. **This histogram 
approximates the sampling distribution for the correlation coefficient**
6. Calculate the bounds such that the middle 95% of the observations are between the
lower and upper bounds. In R, we can do this using `quantile(sample_corr, 0.025)` and
`quantile(sample_corr, 0.975)` to locate the 2.5th and 97.5th percentiles of the 
variable `sample_corr`.

```{r}
# students, you don't need to know how this code works. 

number_of_bootstraps <- 1000
sample_size <- 200

calc_sample_stats <- function(df) {
  df %>% summarize(sample_corr = cor(math, write))
}

set.seed(1)

many_sample_correlations <- replicate(number_of_bootstraps, 
                                      sample_n(hsb2, sample_size, replace = T), 
                                      simplify = F) %>%
  lapply(., calc_sample_stats) %>%
  bind_rows() %>%
  mutate(sample.id = 1:n())

head(many_sample_correlations)
```

### Calculate the lower and upper bounds of the 95% bootstrap CI for the correlation coefficient

```{r}
# Understand this code. It takes the data frame containing the bootstrap
# sample medians and calculates the lower and upper CI using the quantile function.

bounds <- many_sample_correlations %>%
  summarise(lower_CI = quantile(sample_corr, 0.025), 
            upper_CI = quantile(sample_corr, 0.975))

bounds
# how would the code change if we were interested in the 99% CI?
```

### Plot the histogram of the bootstrapped correlations and show the 95% CI

```{r, fig.align="centre", out.width="80%"}
ggplot(many_sample_correlations, aes(x = sample_corr)) + 
  geom_histogram(binwidth = 0.02, col = "white", fill = "#99d8c9") +
  labs(x = "Bootstrapped correlations") + 
  geom_vline(aes(xintercept = corr_math_write %>% pull())) +
  theme_minimal(base_size = 15) +
  geom_vline(aes(xintercept = bounds %>% pull(lower_CI)), lty = 2) +
  geom_vline(aes(xintercept = bounds %>% pull(upper_CI)), lty = 2)
```

### Summary on Bootstrap CIs

- The bootstrap is a method that we use to calculate confidence intervals
- It is particularly useful when:
    - We don't have a nice formula to calculate the CI, or we don't know what the 
    formula is
    - The underlying assumptions of using a "large sample" formula are not satisfied
- We can make bootstrap CIs around **any** statistic we've learnt about: the median,
the quartiles, the correlation coefficient, etc. 
