---
title: 'Lab 6: Introducing the Central Limit Theorem'
output: pdf_document
---

# Introduction 

This is a collaborative assignment that the entire lab section will conduct 
together.  We will all contribute to a central data source (a google sheet) and 
the GSI will summarize the results as the class progresses.  The purpose is to
develop a very concrete idea of sampling distributions, and to see the central 
limit in (live) action.  

# The underlying population

Suppose you had a data frame containing the **entire population** of all
residents of Alameda County. You have data on three variables:

1.  Born either out (=1) versus in (=0) the county.
2.  Number of siblings (integer)
3.  Number of visits to the hospital last year

```{r import-dataset}
library(tidyverse)
alameda_pop <- read_csv("../../Data/L05_Alameda.csv")
head(alameda_pop)
```

1. Read in the data and assign it to the name `alameda_pop`. Calculate the true 
(population) mean, and make histograms or bar charts of the distribution for each 
variable.  

```{r readin, message = F, warning = F}
# write your code here
```

Notice that the distribution of number of siblings and number of hospital visits
are discrete distributions and both skewed right. Today, we will focus on the 
distribution of the number of siblings `num_sibs`. Remember, we know the population
mean exactly, because we have all the data. You calculated the underlying population
mean in the code chunk above.

### Calculating the sampling distribution

We are now going to look at the approximate **sampling distribution for the sample mean**
of the `num_sibs` in (live) action. Remember from earlier lectures that a 
**sampling distribution** is a distribution for a statistic, like the sample 
proportion or the sample mean.

Each student will be tasked with repeatedly taking a random sample of the population. 
Once you take your sample you will compute the sample mean and add it to a 
shared google sheet. As the data is added to the google sheet, notice how the graph 
to the right of the data changes; this illustrate how the sampling distribution varies for 
increasingly larger sample sizes.

The GSIs will provide you the link to the communal google sheet. The columns in 
the sheets are  `n (Sample_size)` `mean(numSibs)` `Name (your sign in)`.

### Your task

1. Randomly generate 10 simple random samples of size $n=5$ from the population.
Calculate the average number of siblings for each of your samples. We wrote code
to start you off, which you can simply copy and paste 10 times to generate ten 
randomly drawn samples and their sample means.

```{r size5}
size_5 <- sample_n(alameda_pop, 5, replace = FALSE)
size_5 %>% summarise(mean = mean(num_sibs))
```

After you've calculated 10 sample means using the above code, copy and paste your
data into the google sheet for your lab section. The links to all the 
google sheets are:

- 101B (Thursday 5-7pm): https://docs.google.com/spreadsheets/d/130FOigRrcdzyMpxSmzxEcdx2TOthUeKZFBpfJgHpbWk/edit?usp=sharing
- 102B (Wednesday 2-4pm): https://docs.google.com/spreadsheets/d/1sVx2Vd_57DWNHRvQ1UP_BhaMuRUnc-H-L-f5Ujuv-fs/edit?usp=sharing
- 103B (Friday 9-11am): https://docs.google.com/spreadsheets/d/1sJpPRIyl83CcFwOyfwJCgyM1LubDZx-RvAfYwG1v8iA/edit?usp=sharing
- 104B (Wednesday 9-11am): https://docs.google.com/spreadsheets/d/1iOUF5bohUL3_tEHkjOjXC_bueFVLm-sA-lWtxHFLgS0/edit?usp=sharing
- 105B (Wednesday 4-6pm):  https://docs.google.com/spreadsheets/d/1PwcGfK4dKbybxFfrwXm8-H4yCzKVBzCenDaf902Btlg/edit?usp=sharing
- 106B (Wednesday 5-7pm): https://docs.google.com/spreadsheets/d/1AP3oewsuzDDpG6oPnMEqXXjzYNY1Q2EBQjycb5aSokg/edit?usp=sharing
- 107B (Thursday 8-10am):  https://docs.google.com/spreadsheets/d/15JdSP4V3-K5BY76sxRXL3wQcDWhNkuAXxPXzqD88Fo0/edit?usp=sharing
- 108B (Wednesday 1-3pm):  https://docs.google.com/spreadsheets/d/151RrhBvC33sb_4mJDl5ibNOPJRKvKQK4LlCCHg-1pDA/edit?usp=sharing
- 109B (Wednesday 9-11am): https://docs.google.com/spreadsheets/d/1bR2eXPNrdKIi_Cxw5nllQFPwR3-KWcvhA93a6IfRxhc/edit?usp=sharing
- 110B (Friday 9-11am):  https://docs.google.com/spreadsheets/d/17Td3U8t4MoH0nCOTevhJjsiENPaYAWT_egRclsxM9IM/edit?usp=sharing

Once the sheet is full, look at the plot of the **sampling distribution** for
the mean number of siblings when $n=5$. 

- What is the range of the sampling distribution of the mean? 

Replace with your answer.

- What is the shape of the sampling distribution of the mean?

Replace with your answer.

- What is the mean of the sampling distribution of the mean?

Replace with your answer.

Once the class has examined the sampling distribution when $n=5$, repeat the 
same steps for n=50. 

2. Repeat for a sample size of $n=50$

```{r size50}
size_50 <- sample_n(alameda_pop, 50, replace = FALSE)
size_50 %>% summarise_all(mean)
```

After you calculated your 10 sample means, navigate to the google sheet from 
before, but switch to the sheet with $n=50$ (you can switch tabs in the bottom left).
Add your data for $n=50$.

Once this is done, look at the plot to the right; now with $n=50$

- What is the range of the sampling distribution of the mean? How does it compare
to when $n=5$?

Replace with your answer.

- What is the shape of the sampling distribution of the mean? How does it compare
to when $n=5$?

Replace with your answer.

- What is the mean of the sampling distribution of the mean?

Replace with your answer.

3. Repeat the entire process for sample size $n=500$.

```{r size500}
size_500 <- sample_n(alameda_pop, 500, replace = FALSE)
size_500 %>% summarise_all(mean)
```

- What is the range of the sampling distribution of the mean? How does it compare
to when $n=5$ and $n=50$?

Replace with your answer.

- What is the shape of the sampling distribution of the mean? How does it compare
to when $n=5$ and $n=50$?

Replace with your answer.

- What is the mean of the sampling distribution of the mean?

Replace with your answer.

4. Suppose you have 5000 classmates, and they have done this lab and added their data to this 
sheet: https://docs.google.com/spreadsheets/d/1AXStOd618raoWvrBequxOh5CDwgisFJoHo50fmcKb_E/edit?usp=sharing

Open the link, and look at the resulting sampling distributions for $n=5$, $n=50$, and $n=500$.
This is what happens when you repeat the sampling 5000 times.
