---
title: 'Lab 7: Confidence intervals for the mean'
output: pdf_document
---

# Introduction 

This is a collaborative assignment that the entire lab section will conduct 
together. We will all contribute to a central data source (a Google sheet) and 
the GSI will summarize the results as the class progresses.  The purpose is to
develop a very concrete idea of confidence intervals by watching their distributions, and to see their performance LIVE!

# The underlying population

Similar to lab 6, suppose you had a data frame containing the **entire population** 
of all residents of Alameda County. You have data on four variables:

1.  Born either out (=1) versus in (=0) the county.
2.  Number of siblings (integer)
3.  Number of visits to the hospital last year
4.  Each individual's height (inches)

1. Read in the data and assign it to the name `alameda_pop`. Calculate the true 
(population) mean of the variable `height`, and the population standard deviation.
Make a histogram of `height` and comment its distribution. Use ggplot() to quickly
make a ggplot. Does it look Normally distributed?

```{r, message = F, warning = F}
library(tidyverse)
# write your code here
```

<Answer the questions here.>

Suppose that you know the population standard deviation. It is what you calculated
in the previous code chunk. Use the following code chunk to save its value. Replace
the word `NULL` with your calculation of the population standard deviation.:

```{r save-known-pop-sd}
known_pop_sd <- NULL
```

Now, suppose you *do not know* the population mean, but wanted to estimate it 
based on a sample. In this lab, we actually know the true value because we 
calculated it above. This way, we can see how well any one sample does at estimating
the population mean and see how often the confidence intervals contain the mean
across several repeated samples.

### Calculating the CI and looking at its performance

We are now going to compute (and enter into our master google sheet) 95%
confidence intervals (CI) for sampling means of different sizes. For this lab we:

- Have a variable with an underlying Normal distribution
- Will take simple random samples (SRS) from this distribution
- Know the value of the population mean (from your calculation in the first code
chunk)

Thus, we satisfy the three conditions discussed in Wednesday's lecture for calculating
a confidence interval when the underlying SD is known.

Recall the format for the 95% confidence interval in this setting (Hover with 
your mouse within the double-dollar signs to see the formula or knit the file to 
read them easily):

$$\bar{x} \pm 1.96 \times\frac{\sigma}{\sqrt{n}}$$
Where:

- The sample mean is the estimated mean, based on your sample
- $\sigma$ is the known standard deviation `known_pop_sd`, that you saved earlier for the distribution of `height`
- $\sigma/\sqrt{n}$ is the standard error of the sampling distribution for $\bar{x}$
- 1.96 is the critical value for a 95% CI.

Like last lab, each student will take a few samples from the distribution. This
time, you will calculate the mean of your sampled heights and its confidence
interval using the above formula. We will then record this information into the 
google sheet and plot all the CIs when we have at least 20 of them.

Here are the links to the communal Google sheets. Open the one for your lab.
The columns in the sheets are `sample_mean_heights` `lower_CI` `upper_CI` 
`sample_size` `student`.

Lab 101B: https://docs.google.com/spreadsheets/d/1bWatXMlpJLvOR0ch4Hox-5kMHcsit-32nAuPXhYgXMI/edit?usp=sharing
Lab 102B: https://docs.google.com/spreadsheets/d/1V8TtSV6uv7prZIYcGQTn9QQ482IajVYLX6E4LZjwfRw/edit?usp=sharing
Lab 103B: https://docs.google.com/spreadsheets/d/1diY17Q78CVczyLFOSwlySRwF0Ef9T96o8IYO6lHEjPA/edit?usp=sharing
Lab 104B: https://docs.google.com/spreadsheets/d/1gVGxDOPrlk3jvKtI5raezOj17WF1IhJaxt3gs1KR4Ds/edit?usp=sharing
Lab 105B: https://docs.google.com/spreadsheets/d/1sYpg1MkTcIR2ViHdQZrtVJ7_oR5OUwVJT-f_LwmBjQo/edit?usp=sharing
Lab 106B: https://docs.google.com/spreadsheets/d/1TMpkxrOXvfp7-u6KxnQiIRFvUuIdrFDNDl6hHwSqtS0/edit?usp=sharing
Lab 107B: https://docs.google.com/spreadsheets/d/15V6qR37-pPW0t3DAAuH_ElyqToTBgtFp0DYLu3iR0Og/edit?usp=sharing
Lab 108B: https://docs.google.com/spreadsheets/d/1TFQYLia1HG1GA6hPGy1WwoFYS-qav5fvcbshwrqD7zs/edit?usp=sharing
Lab 109B: https://docs.google.com/spreadsheets/d/1Axn4186utxbIUGcKu_SNp_u2pd7rtTPNsSlv4HTmvps/edit?usp=sharing
Lab 110B: https://docs.google.com/spreadsheets/d/1pwv2WJbpy_oZZT7D5TN0Gb8-vh-sFx5SkK60Fkr51AI/edit?usp=sharing

### Your task

1. Randomly generate 3 simple random samples of size $n = 10$ from the population.
Calculate the average number of siblings for each of your samples. We wrote code
to start you off, **but you need to replace the three instances of NULL with 
calculations** to compute the sample mean (`sample_mean_heights`), the lower 
confidence interval (`lower_CI`) and the upper confidence interval (`upper_CI`).
Hint: Review the above section for tips on how to calculate the CI if you forget.
Once you do this, you can simply copy and paste 3 times to generate three 
randomly-drawn samples and their sample means.

```{r sample-size-10}
sample_size <- 10
critical_value <- 1.96

size_10 <- sample_n(alameda_pop, sample_size, replace = FALSE)
size_10 %>% summarise(mean_heights = NULL) %>%
mutate(lower_CI = NULL,
       upper_CI = NULL
)
```

Navigate to the google sheet for your lab and add the mean and its CI for your
three samples. Once this is done enough times, the GSI can make a plot of the
CIs and see how many contain the true value for the mean. Based on this plot:

- What proportion of the confidence intervals contain the mean? 

Type your answer here.

2. Repeat for a sample size of $n=50$. In the code chunk below, generalize your
code from the previous chunk to create three samples, this time of size 50:

```{r sample-size-50}
# add in your code here.
```

After you calculated your 3 sample means, navigate to the Google sheet from 
before, but switch to the sheet with $n=50$ (you can switch tabs in the bottom left).
Add your data for $n=50$.

Once this is done, the GSI will plot these data, now with $n=50$

- What proportion of the confidence intervals contain the mean?

Type your answer here.

- How do the average widths of the CI's compare for $n=50$ versus $n=10$?

Type your answer here.

- What would happen to the confidence intervals if $n=500$?

Type your answer here.

**Submission** 

Please submit this lab file *directly* to Gradescope under the "Lab 6/7" 
Assignment (You don't need to submit the CLT lab file). You can do this by 
knitting your file and downloading the PDF to your computer. Then navigate to 
Gradescope.com to submit your assignment. Here is a tutorial if you need help: https://www.gradescope.com/get_started. Scroll down on that page to 
"For students:submitting homework".
