---
title: 'Lab 08: Proportions and Inference'
subtitle: "Solutions"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

### Tests of changes in sex ratios based on a single sample

There is a long literature studying changes in sex-ratios of births due to 
stressful events, such as 9/11. In today's lab, we consider a relatively small 
study that recorded biomarkers of stress on pregnancy.  In the group of subjects
that had the highest markers of stress (based on cortisol), there were 14 births
to males out of a total of 38.

In this lab, we will compare the four methods we learned to calculate CIs for
proportions. Recall that two of these methods involved hand calculations (though
we can treat R as if it were a calculator) and two of the methods used built-in
R functions.

1. Use the Normal approximation to construct a 95% confidence interval
in this high stress group. We also called this specific method of 
constructing the CI the "large sample method"

```{r large-sample-method}
# your code here

# solution
p.hat <- 14/38 # estimate proportion
se <- sqrt(p.hat*(1-p.hat)/38) # standard error
c(p.hat - 1.96*se, p.hat + 1.96*se) # CI
```

Solution: The 95% CI goes from 21.5% to 52.2% using the large sample method.

2. Create the 95% CI again, this time using the R function that implements the 
Wilson Score method with a continuity correction.

```{r wilson-score}
# your code here

# solution
prop.test(14, 38, p=0.5)
```

Solution: The 95% CI goes from 22.3% to 54.0% using the Wilson Score method with 
continuity correction.

3.  Create the 95% CI again, this time using the Plus 4 method.

```{r plus-four}
# your code here

# solution
p.tilde <- (14 + 2)/(38 + 4)
se <- sqrt(p.tilde*(1 - p.tilde)/42) # standard error
c(p.tilde - 1.96*se, p.tilde + 1.96*se) # CI
```

Solution: The 95% CI goes from 23.4% to 52.8% using the Plus 4 method.

4.  Create the 95% CI again, this time using the R function that implements the 
Clopper Pearson (Exact) method.

```{r exact-method}
# your code here

# solution
binom.test(14, 38, p=0.5)
```

Solution: The 95% CI goes from 21.8% to 54.0% using the Exact method.

5. Summarize the four methods' estimates in the following table. Do they include 
the null hypothesized value for the sex ratio?

| Method        | 95% Confidence Interval | 
|---------------|:-----------------------:|
| Large sample  | AA.A% to AA.A%          |
| Wilson Score* | AA.A% to AA.A%          | 
| Plus four     | AA.A% to AA.A%          | 
| Exact         | AA.A% to AA.A%          | 

* with continuity correction.

Solution:

| Method        | 95% Confidence Interval | 
|---------------|:-----------------------:|
| Large sample  | 21.5% to 52.2%          |
| Wilson Score* | 22.3% to 54.0%          | 
| Plus four     | 23.4% to 52.8%          | 
| Exact         | 21.8% to 54.0%          | 

* with continuity correction.

6. Here is a code template to help you to graphically present these estimates. 
Graphical presentations of estimates and their CIs is very useful for assessing
whether the CIs overlap the null hypothesized value and tends to be better than
presenting tables of estimates to readers of your research.

```{r}
# First make a tibble (an easy way to make a data frame) with the data about
# each confidence interval. To do this, replace each instance of 0.00 with the
# estimate from your calculations above.

library(ggplot2)
library(tibble)
sex_CIs <- tibble(method   = c("large sample", "Exact", "Wilson", "Plus 4"),
                  lower_CI = c(0.0           , 0.0    ,  0.0    ,  0.0),
                  upper_CI = c(0.0           , 0.0    ,  0.0    ,  0.0),
                  estimate = c(0.0           , 0.0    ,  0.0    ,  0.0)
                  )

# Build the ggplot incrementally, to understand how it works.
# Step 1: (qu: why do we put a horizontal line at 50?)
ggplot(data = sex_CIs, aes(x = method, y = estimate)) + 
  geom_point() + 
  geom_hline(aes(yintercept = 50), lty = 2)

# Step 2:
ggplot(data = sex_CIs, aes(x = method, y = estimate)) + 
  geom_point() +
  geom_hline(aes(yintercept = 50), lty = 2) +
  geom_segment(aes(x = method, xend = method, y = lower_CI, yend = upper_CI)) +
  labs(y = "Estimate with 95% CI")

# Solution 

sex_CIs <- tibble(method   = c("large sample", "Exact", "Wilson", "Plus 4"),
                  lower_CI = c(21.5          , 21.8   ,  22.3   ,  23.4),
                  upper_CI = c(52.2          , 54.0   ,  54.0   ,  52.8),
                  estimate = c(36.8          , 36.8   ,  36.8   ,  36.8)
                  )

ggplot(data = sex_CIs, aes(x = method, y = estimate)) + 
  geom_point() +
  geom_hline(aes(yintercept = 50), lty = 2) +
  geom_segment(aes(x = method, xend = method, y = lower_CI, yend = upper_CI)) +
  labs(y = "Estimate with 95% CI")

```

What does `geom_segment()` do? In particular, what do `x`, `xend`, `y` and `yend`
specify in this case?

Solution: `geom_segment()` creates a segment of a line that goes between the
coordinate denoted by (`x`, `y`) the the one denoted by (`xend`, `yend`).

7. Based on this plot, what can you say about the confidence intervals for the 
sex ratio in the high stress group?

The estimated proportion of male birts is 36.8%. Because all the CIs overlap the 
null value of 50%, the null hypothesis of $p_0 = 50\%$ is not rejected at a 5%
level. However, most of the values in the CI are <50%, and might suggest that 
with more data the null hypothesis could be rejected (as the CIs become narrower
as sample size increases.)

--------------------------------------------------------------------------------

8. If you have time, repeat the above analysis for the group with low stress. There
were 25 births to this group, of which 17 of them were to males.

```{r}
# your code here

# solution

#large sample
p.hat <- 17/25 # estimate proportion
se <- sqrt(p.hat*(1-p.hat)/25) # standard error
c(p.hat - 1.96*se, p.hat + 1.96*se) # CI

#exact
binom.test(17, 25, p=0.5)

#wilson score
prop.test(17, 25, p=0.5)

#plus four
p.tilde <- (17 + 2)/(25 + 4)
se <- sqrt(p.tilde*(1 - p.tilde)/42) # standard error
c(p.tilde - 1.96*se, p.tilde + 1.96*se) # CI

sex_CIs <- tibble(method   = c("large sample", "Exact", "Wilson", "Plus 4"),
                  lower_CI = c(49.7          , 46.5   ,  46.4   ,  51.1),
                  upper_CI = c(86.2          , 85.1   ,  84.3   ,  79.9),
                  estimate = c(68.0          , 68.0   ,  68.0   ,  68.0)
                  )

ggplot(data = sex_CIs, aes(x = method, y = estimate)) + 
  geom_point() +
  geom_hline(aes(yintercept = 50), lty = 2) +
  geom_segment(aes(x = method, xend = method, y = lower_CI, yend = upper_CI)) +
  labs(y = "Estimate with 95% CI")
```

7. If you recreated the graph for the low stress group, what can you say about 
the confidence intervals for the sex ratio in this group?

Solution: 
The estimated proportion of male birts is 68.0% in this group. Three of the CIs
overlap the null hypothesized value of 50%, while the plus 4 method does not. 
Most of the values contained in the CIs are >50%, and might suggest that with 
more data the null hypothesis could be rejected (as the CIs become narrower
as sample size increases.) Overall the CI are so wide that they are not very useful,
and more data would be helping in deciding if the sex ratio is affected by stress.



**Submission** 

Please submit this lab file *directly* to Gradescope under the "Lab 8" 
Assignment. You can do this by 
knitting your file and downloading the PDF to your computer. Then navigate to 
Gradescope.com to submit your assignment. Here is a tutorial if you need help: https://www.gradescope.com/get_started. Scroll down on that page to 
"For students:submitting homework".
