---
title: "Alternative Estimates of Correlation"
date: "11/23/2019"
output: pdf_document
---

This supplement to your housing price lab includes two practice questions 
regarding nonparametric estimation of the correlation coefficient.

We are still using the Boston housing data set.  However, we will be loading a 
different version of the data that includes granular geographic variables.

First, make sure the following packages are loaded:

```{r, warning = FALSE, message = FALSE}

### the classics
library(ggplot2)
library(tidyr)
library(broom)
library(dplyr)
```

Now, load the Boston housing data with new variables:
```{r}
housing_data <- readRDS("../../Data/boston_housing_data_with_geo_vars.RDS")
```

**Question 1**

I am interested in testing whether median home value in 1000s (`MEDV`) 
is correlated with pupil teacher ratio (`PTRATIO`) by town. Take a look at the 
data and see that both of these variables are continuous. We would like to test
the null hypothesis that the correlation coefficient equal to 0.  

Using methods covered in the textbook, we learned that if we were to reject the 
hypothesis that the slope coefficient was equal to 0 for the relationship between 
`MEDV` and `PTRATIO`, then this would imply that we would also reject the hypothesis
that the correlation coefficient is equal to 0. The reason this is true is that
the formula for the slope parameter is a function of the correlation coefficient:

$$\hat{b} = r * \frac{sd_y}{sd_x}$$ where $r$ is the 
correlation between $x$ and $y$). 

**1(a) Name another method I can use to test whether there is a nonzero correlation between MEDV and PTRATIO.**

Your answer here.

**1(b) Why might we prefer this method to the regression-based test?**

Your answer here.

**1(c) Describe in words how you might carry out this alternative test.**

Your answer here.

**1(d) [Optional] Run this alternative test using the housing data.**

Your code and answer here

\newpage


**Question 2**

Now, I am interested in creating a 95% confidence interval 
for the correlation between median home value in 1000s (`MEDV`) 
and pupil teacher ratio (`PTRATIO`) by town.

Think back to the formulas we covered in class (from the textbook) to create
confidence intervals. None of those formulas applied to the correlation coefficient.
However, we learned an alternative method to calculate confidence intervals.

**2(a) What alternative method can I use to create a 95% confidence interval for the correlation between `MEDV` and `PTRATIO`?**

Your answer here.

**2(b) Why might I prefer using this method to the t-distribution based methods?**

Your answer here.

**2(c) Describe in words how you might carry out this alternative method.**

Your answer here.

**2(d) [Optional] Code this alternative method for our data in R.**

Your code and answer here.


