---
title: "Alternative Estimates of Correlation"
date: "11/23/2019"
output: pdf_document
---

This supplement to your housing price lab includes two practice questions 
regarding nonparametric estimation of the correlation coefficient.

We are still using the Boston housing data set.  However, we will be loading a 
different version of the data that includes granular geographic variables.

First, make sure the following packages are loaded:

```{r, warning = FALSE, message = FALSE}

### the classics
library(ggplot2)
library(tidyr)
library(broom)
library(dplyr)
```

Now, we will load the Boston housing data with new variables:
```{r}
housing_data <- 
  readRDS("../../Data/boston_housing_data_with_geo_vars.RDS")
```


**Question 1**

I am interested in testing whether median home value in 1000s (`MEDV`) 
is correlated with pupil teacher ratio (`PTRATIO`) by town. Take a look at the 
data and see that both of these variables are continuous. We would like to test
the null hypothesis that the correlation coefficient equal to 0.  

Using methods covered in the textbook, we learned that if we were to reject the 
hypothesis that the slope coefficient was equal to 0 for the relationship between 
`MEDV` and `PTRATIO`, then this would imply that we would also reject the hypothesis
that the correlation coefficient is equal to 0. The reason this is true is that
the formula for the slope parameter is a function of the correlation coefficient:

$$\hat{b} = r * \frac{sd_y}{sd_x}$$ where $r$ is the 
correlation between $x$ and $y$). 

**1(a) Name another method I can use to test whether there is a nonzero correlation between MEDV and PTRATIO.**

SOLUTION: Permutation test.

**1(b) Why might we prefer this method to either the regression-based test or a t-test?**

SOLUTION: The permutation test does not require that our sample data be 
generated from an approximately normal distribution.

**1(c) Describe in words how you might carry out this nonparametric test.**

SOLUTION: The procedure would go as follows:

i. Calculate the sample correlation.
ii. While keeping the order of PTRATIO fixed, randomly shuffle the values of MEDV.
iii. Calculate and store the estimate of correlation between PTRATIO and MEDV based on the shuffled data.
iv. Iterate steps (ii) and (iii) preferably at least 10,000 times.
v. Plot a histogram of these 10,000 estimated correlations.
vi. We have created an approximated distribution of possible sample correlations under the null hypothesis that PTRATIO and MEDV are unrelated.  Calculate the 97.5th (or 2.5th if sample correlation 
is negative) quantile of our approximated distribution.
vii. Reject the null hypothesis if the sample correlation is 
greater (or less) than the quantile calcualted in step (vi).


**1(d) [Optional] Run this nonparametric test for our data in R. Note that you do 
not need to know how to write the code for the exam**

SOLUTION:

```{r}
### set seed for replicability
set.seed(20191123)

### compute actual observed correlation
sample_correlation <-
  cor(housing_data$PTRATIO, housing_data$MEDV)
### print
sample_correlation

### fix order of ptratio vector
fixed_ptratio <- housing_data$PTRATIO

### storage for randomized permutations
synthetic_distribution <-
  ### just creates 10,000 filler 0's for now
  rep(0, 10000)

for(i in 1:10000) {
  ### randomly arrange medv
  rearranged_medv <- sample(housing_data$MEDV)
  
  ### NOTE: normally we would just compute every possible
  ### randomization but that is computationally expensive
  ### for a large dataset.
  
  ### compute randomized correlation and store
  synthetic_distribution[i] <- 
    cor(fixed_ptratio, rearranged_medv)
  
}

### visualize synthetic distribution
### convert to data frame form in order to use ggplot
synthetic_distribution_df <-
  data.frame(synthetic_distribution)

synthetic_distribution_df %>%
  ggplot(aes(x = synthetic_distribution)) + geom_histogram()

### calculate 2.5th quantile since we have negative sample correlation
quantile(synthetic_distribution, .025)

### Do we reject the null hypothesis?
sample_correlation < quantile(synthetic_distribution, .025)

```

We reject the null hypothesis that there is no relationship 
between neighborhood pupil-teacher ratio and median home value. 

\newpage

**Question 2**

Now, I am interested in creating a 95% confidence interval 
for the correlation between median home value in 1000s (`MEDV`) 
and pupil teacher ratio (`PTRATIO`) by town.

Think back to the formulas we covered in class (from the textbook) to create
confidence intervals. None of those formulas applied to the correlation coefficient.
However, we learned an alternative method to calculate confidence intervals.

**2(a) What alternative method can I use to create a 95% confidence interval for the correlation between `MEDV` and `PTRATIO`?**

SOLUTION: Bootstrap

**2(b) Why might I prefer using this method to the t-distribution based methods?**

SOLUTION: My data do not approximately follow the normal distribution.

**2(c) Describe in words how you might carry out this alternative method.**

SOLUTION: My procedure would be as follows:

i. With replacement, sample 506 **pairs** of PTRATIO and MEDV 
from the data.  Since these variables are paired, it is important that 
we do not separately sample PTRATIO and then sample MEDV.

ii. Compute the correlation between PTRATIO and MEDV for 
this synthetic sample of our data.

iii. Repeat steps (i) and (ii) enough times to create a good 
synthetic distribution for the correlation (preferably 10,000 or more 
times).

iv. Plot a histogram of the bootstrap.

v. Calculate the 2.5th and 97.5th quantiles of our synthetic 
distribution.  These are the bounds of our bootstrap confidence 
interval for the correlation between MEDV and PTRATIO.

**2(d) [Optional] Code this nonparametric method for our data in R.**

SOLUTION:

```{r}
### set seed for replicability
set.seed(20191123)

### create storage vector for bootstrap correlations
bootstrap_corrs <-
  ### just creates 10,000 filler 0's for now
  rep(0, 10000)

for(i in 1:10000) {
  ### take paired sample of data
  bootstrap_sample <-
    ### dplyr sample_n function samples rows at a time.
    sample_n(housing_data %>%
               select(MEDV, PTRATIO),
           506,
           replace = TRUE)
  
  ### fill in the boostrap correlation vector
  bootstrap_corrs[i] <-
    cor(bootstrap_sample$MEDV,
        bootstrap_sample$PTRATIO)
  
}

### plot our synthetic distribution
### convert to data frame form in order to use ggplot
bootstrap_corrs_df <-
  data.frame(bootstrap_corrs)

bootstrap_corrs_df %>%
  ggplot(aes(x = bootstrap_corrs)) +
  geom_histogram()

### compute the CI bounds using the quantile function.
ci_lower <- 
  quantile(bootstrap_corrs, 0.025)

ci_upper <- 
  quantile(bootstrap_corrs, 0.975)

### report the CI
c(ci_lower, ci_upper)


```



