---
title: "Spatial Visualization Tutorial"
date: "11/23/2019"
# note to add the following line so R knows to create an interactive document
runtime: shiny
output: html_document
---


This optional supplement to your housing price lab includes:

\item __I__ An introduction to spatial visualization of data (just for your interest, no need to study)

We are still using the Boston housing data set.  However, we will be loading a 
different version of the data that includes granular geographic variables.

First, make sure the following packages are installed:
```{r, warning = FALSE, message = FALSE}
### packages we will use:
### (comment-out runs of install.packages out once installed)

### helpful HTML functions
#install.packages("httpuv)
library(httpuv)
#install.packages("htmltools)
library(htmltools)
### R's web app creation package
library(shiny)

### Our Mapping Package
#install.packages("leaflet")
library(leaflet)
### only installs tigris package if it has not yet been installed.
if(!"tigris" %in% installed.packages()) {
  install.packages("tigris") } 

library(tigris)
#install.packages("sf")
library(sf)


### Helpful coloring functions
#install.packages("RColorBrewer")
library(RColorBrewer)

### the classics
library(ggplot2)
library(tidyr)
library(broom)
library(dplyr)
```

Now, we will load the Boston housing data with new variables:
```{r}
housing_data <- 
  readRDS("../../Data/boston_housing_data_with_geo_vars.RDS")
```


# Bonus: Interactive Spatial Plots and For-Report Spatial Plots

Oftentimes, especially during the exploratory stage of research, 
investigators are interested in examining possible spatial 
relationships of data.

Our Boston housing data presents an especially good opportunity 
to see if there are any interesting spatial patterns in our 
variables.

The Leaflet package from R uses HTML code to create interactice maps. 
So, when we send the interactive Leaflet maps to someone reviewing our work, they can zoom, hover over data labels, and 
enjoy other interactive functions that we program into the map. 


First, I will create a heatmap of median neighborhood home values 
using Leaflet.

One thing to keep in mind is that for almost all of this code, 
I googled how to do things and then slightly altered code snippets 
that were available online.  If you don't know how to do something, 
chances are you can find a helpful answer on the web.

```{r}
### I want to color my heatmap data from yellow to red according to 
### the home value. 
### I googled for this code snippet to color scale a continuous variable
pal <- colorNumeric(
  palette = colorRampPalette(c('yellow', 'red'))(length(housing_data$MEDV)), 
  domain = housing_data$MEDV)


### These functions are from the Leaflet package. almost all of this I made by
### googling how to do so.
housing_data %>%
  leaflet() %>%
  addTiles() %>%
  addCircleMarkers(lng = ~lon, lat = ~lat,
                   color = ~pal(MEDV),
                   popup = ~as.character(MEDV),
                   radius = 0.5) %>%
  ### This is the background art for the map.
  addProviderTiles(providers$CartoDB.Positron) %>%
  addLegend(position = "topright", pal = pal, 
            values = housing_data$MEDV, bins = 7,
            ### HTML function allows me to add line break using <br/>
            title = htmltools::HTML("Median Home <br/> Value ($1,000's)")) %>%
  addControl("Boston Home Values, 1978", position = "bottomleft")


```

We can take the interactivity of the map one step further.  R has 
developed tools that allow us to build web apps (through the "Shiny" 
package).

Using R's Shiny tools, we can create a drop down list of all of our housing data variables and allow the user 
to select whichever map they are interested in.


```{r}
### Initial data cleaning
### I create a list of variable descriptions for my legends.
### Note: "<br/>" is HTML syntax for a line break since Leaflet plots and shiny use HTML language (as we are building a web app).
legend_desc <-
  list(
    "Median Home <br/> Value ($1,000's)",
    "Per Capita Crime Rate",
    "Prop. of Land Zoned <br/> Lots of >25000 sq. ft.",
    "Prop. of Non-Retail Business Acres",
    "Charles River Dummy",
    "Nitrous Oxide Concentration <br/> PP10Mil",
    "Avg. Num. Rooms per Dwelling",
    "Prop. of Building Constructed <br/> Prior to 1940",
    "Avg Distance to 5 Employment Centers",
    "Radial Highway <br/> Accessibility Index",
    "Property-Tax Rate",
    "Pupil-Teacher Ratio",
    "Prop. Lower Status Population"
  )
names(legend_desc) <- 
  names(housing_data[1:length(legend_desc)])


```

```{r}

shinyApp(
  ui = 
  fluidPage(
    ### tells Shiny interface that you are printing a Leaflet Map
    leafletOutput("my_map"),
    ### creates variable drop-down list in my web app
    selectInput(### How I will reference the option in subsequent code:
                "variable",
                ### Dropdown text that will appear in the app:
                "Variable:",
                names(housing_data %>%
                        dplyr::select(MEDV:LSTAT,-CHAS)),
                ### I choose the initial selected option for the map:
                selected = "MEDV"),
    ### I create a drop-down list for heatmap coloring options
    selectInput("colors", "Color Scheme",
                ### list of common heatmap coloring schemes I found online
                rownames(subset(brewer.pal.info, 
                                category %in% c("seq", "div"))),
                ### default option
                selected = "YlOrRd"
                ) 
  ),
  server = function(input, output) {
    ### first create non-dynamic elements of the map
    my_map = housing_data %>%
      leaflet() %>%
      addTiles() %>%
      ### sets a default dimension of the map
      fitBounds(~min(lon), ~min(lat), ~max(lon), ~max(lat))
  
  # The reactive function notes this input for the map will change as the user makes selections in the web app.
  my_pal <- reactive({
    colorNumeric(
      palette = input$colors,
      domain = housing_data[, input$variable])
  })  
  
  
  
  # Incremental changes to the map (in this case, replacing the
  # circles when a new color is chosen) should be performed in
  # an observer. Again, I found all of this code online.
  observe({
    pal <- my_pal()
    
    leafletProxy("my_map", data = housing_data) %>%
      ### remove any existing layers of data
      clearMarkers() %>%
      clearControls() %>%
      
      addCircleMarkers(
        lng = ~ lon,
        lat = ~ lat,
        color = as.formula(paste0("~pal(", input$variable, ")")),
        popup = as.formula(paste0("~as.character(", input$variable, ")")),
        radius = 0.5
      ) %>%
      ### This is the background art for the map.
      addProviderTiles(providers$CartoDB.Positron) %>%
      addLegend(
        position = "topright",
        pal = pal,
        ### I run "as.formula" since my input calls a character string of the selected variable.
        values = as.formula(paste0("~", input$variable)),
        bins = 7,
        # HTML function allows me to add line break using <br/>
        title = htmltools::HTML(as.character(legend_desc[input$variable]))
      ) %>%
      addControl(HTML("<b>Boston Housing Data by Neighborhood, 1978"), position = "bottomleft")
  })
    ### Tells Shiny to render the map I created above.
    output$my_map <- renderLeaflet(my_map)
  },
  ### controls the dimensions the app will take in the Web Document
  options = list(height = 750)
)

```



While the interactive maps are great, especially in the exploratory stage, it may be preferable 
to create a static map with ggplot formatting for a final research report.  We can do this using 
the SF package that we loaded earlier.

Our Boston housing data has a 'geometry' variable, which includes shape data for each neighborhood 
of interest.  The shape data provides boundary coordinates for each neighborhood, which R can utilize 
to draw a map of each tract of land.


```{r, results = "hide"}
### first, I pull generic state shape data from the TIGRIS package, which has a lot of useful 
###   shape files for us.

### We could try to look at the results drawn over a map of Massachussetts

### I am currently using the sf package to draw maps, so this is the class I draw.
mass_geo <- tigris::counties(class = "sf", cb = TRUE) %>%
  ### filter data to massachussetts
  filter(STATEFP == "25")
```

```{r}
### Take a look at the geometry variables.  They are special types of objects that 
###  list out geographical coordinates that R should draw lines between to create 
###   shapes of states, counties, etc.

### First heatmap plot overlaid over MA
ggplot() +
  ### our housing data.
  geom_sf(data = housing_data,
          aes(geometry = geometry,
              fill = MEDV))  +
  ### add MA background
  geom_sf(data = mass_geo)

```

It's difficult to see what's going on, but we could see how this would be a visually 
pleasing layer to add to the map if we had more Massachussetts data than just Boston.

I will remove the state layer for now and just look at our Boston data.

```{r}
ggplot() +
  ### our housing data.
  geom_sf(data = housing_data,
          aes(geometry = geometry,
              fill = MEDV)) +
  ### I want to go back to yellow-red coloring as well
  scale_fill_gradient(low = "yellow", high = "red") +
  ggtitle("Median Neighborhood Home Values\n Boston Metropolitan Area, 1978")

      
```


There is still a lot of work to be done to make this plot look better.  But this is an introduction to 
creating static maps that you might place in a final research report.


