---
title: "Assignment 2: Summarize global cesarean delivery rates and GDP across 137 countries"
author: "Your name and student ID"
date: "Today's date"
output: pdf_document
---

* Due date: Friday, September 13 5:00pm.
* Late penalty: 50% late penalty if submitted within 24 hours of due date, no 
marks for assignments submitted thereafter.
* This assignment is marked out of 27. Marks are indicated for each 
question. There are 15 questions total.
* Submission process: Unlike previous labs and assignments, please submit your 
assignment *directly* to Gradescope. You can do this by knitting your file and 
downloading the PDF to your computer. Then navigate to Gradescope.com to submit 
your assignment. Here is a tutorial if you need help: https://www.gradescope.com/get_started. 
Scroll down on that page to "For students:submitting homework".

Helpful hints:

- Every function you need to use was taught during lecture! So you may need to 
revisit the lecture code to help you along by opening the relevant files on data 
hub. Alternatively, you may wish to view the code in the condensed PDFs posted 
on bCourses (under Files > Lectures). Good luck!

- Knit your file early and often to minimize knitting errors! If you copy and 
paste code for the slides, you are bound to get an error that is hard to 
diagnose. Typing out the code is the way to smooth knitting! We recommend 
knitting your file each time after you write a few sentences/add a new code 
chunk, so you can detect the source of knitting errors more easily. This will 
save you and the GSIs from frustration!

- If your code runs off the page of the knitted PDF then you will LOSE POINTS! To
avoid this, have a look at your knitted PDF and ensure all the code fits in the 
file. When it doesn't, go back to your .Rmd file and add spaces (new lines) using
the return or enter key so that the code runs onto the next line.
 
--------------------------------------------------------------------------------
\newpage

# Summarizing global cesarean delivery rates and GDP across 137 countries

### Introduction 

Recall from this week's lab that we constructed bar charts and histograms to 
explore a data set that looked at global rates of cesarean delivery and GDP. 
If you need a refresher, you can view your knitted file from lab and remind 
yourself what you found. 

In this week's assignment, you will describe these distributions using numbers. 
You will investigate the **mean** and **median** of the distribution of GDP. 
You will also examine the distribution of cesarean delivery separately for 
countries of varying income levels. Lastly, you will describe the **spread**
of the distributions using **quartiles** and make a **box plot**. 

Execute this code chunk to load the required libraries:
```{r load-libraries}
library(readr)
library(dplyr)
library(ggplot2)
```

Just like last time, read in the data that is stored as a .csv file and assign 
it to an object called `CS_data`. We will also use dplyr's `mutate()` to create
the new cesarean delivery variable that ranges between 0 and 100: 

```{r read-in-the-data}
CS_data <- read_csv("./Data/Cesarean.csv")

# this code reorders the Factor variable `Income_Group` in the 
# order specified in this function. This will affect the order the ggplot
# panels are shown in question 8 when we use `facet_wrap()`.
CS_data$Income_Group <- forcats::fct_relevel(CS_data$Income_Group, 
                                            "Low income", "Lower middle income", 
                                            "Upper middle income", "High income: nonOECD", 
                                            "High income: OECD")

CS_data_new <- CS_data %>% mutate(CS_rate_100 = CS_rate*100)
```

\newpage

**1. [1.5 marks] Fill in the blanks indicated by "---" by saving the answer to each blank in the code chunk below:**

Solution: The function `mutate()` takes the old variable called **CS_rate** and multiplies it
by **100** to make a new variable called **CS_rate_100**.
\newpage

**Investigate what would have happened had we not assigned the data using `<-` to 
`CS_data_new`? Re-run the code without the assignment and see what happens.**

```{r no-assignment-example}
rm(CS_data_new) #first, remove using rm() CS_data_new from the environment
CS_data %>% mutate(CS_rate_100 = CS_rate*100) #re-run code without assignment
names(CS_data)
```

Did `CS_rate_100` get added to the data set? No. You can tell by using 
`head(CS_data)` to view the first few rows and notice that the variable hasn't 
been added. This is because when we don't assign the output to anything, 
it just prints it out for us to see. Nothing is saved. Because we are only making
a small change, we can overwrite the original object (CS_data) by assigning the 
transformed data to the same name. Otherwise, you would want to assign the 
variable to a different name so you can keep two separate copies of the data set.

```{r overwrite-original-data}
CS_data <- CS_data %>% mutate(CS_rate_100 = CS_rate*100)
```

\newpage

### GDP: Summarizing measures of centrality

Recall your histogram of GDP from this week's lab:
```{r histogram-GDP-2006}
ggplot(data = CS_data, aes(x = GDP_2006)) +
  geom_histogram(col = "white", fill = "sienna2", binwidth = 1000) + 
  xlab("GDP in 2006") +
  theme_minimal() 
```
\newpage

**2. [1 mark] Describe the shape of this distribution. Is it "skewed left", "skewed right", "symmetric", or "bimodal"?**

[1 mark] Solution: skewed right.
\newpage

**3. [1 mark] Based on your answer, will the mean be approximately the "same", "larger than", or "smaller than" the median?**

[1 mark] Solution: Mean will be larger than the median.
\newpage

**4. [3 marks, not auto-marked] Describe, in words, how the mean and median are calculated:**

Solution:
[3 marks total]
   - [1 mark]: Mean: add up all the measurement values and divide by their total observation count.
   - [2 mark]: Median: First, order the measurements by their value. 
   If the total observation count is an odd number, 
   the middle observation is the median. If an even number, add the
   two mid measurements values and divide by two to calculate the median.

\newpage

To calculate the mean and median in R, we can use the `summarize()` function 
from the `dplyr` package. The `summarize()` function is used anytime we want to 
take a variable and summarize something about it into one number, like it's mean 
or median. Here is the code to summarize `GDP_2006`'s mean and print it out to 
the screen. In the code, we name the mean `mean_GDP` and output the result to 
the screen:

```{r summarize-mean}
GDP_summary <- CS_data %>% summarize(mean_GDP = mean(GDP_2006))
GDP_summary
```
\newpage

**5. [1 mark] Extend the above code to also summarize the median. Call the median 
summary `median_GDP`. Assign this summary to `GDP_summary` (it will overwrite the 
previous version):**

```{r summarize-median}
#solution [1 mark]
GDP_summary <- CS_data %>% summarize(mean_GDP = mean(GDP_2006), 
                                     median_GDP = median(GDP_2006))
GDP_summary
```
\newpage

**6. [2 marks, not auto-marked] `geom_vline()` can be used to add the mean and the median to the 
histogram shown above. This `geom` adds a vertical line to the graph. You need 
to specify where to add the line by passing it an "xintercept" argument. Remove 
the comments (i.e., the three "#") from the code below and update the 
`geom_vline()` code to plot lines at the mean and median by telling it the mean 
and median estimates. The argument `lty=1` (standing for line type) will plot a 
solid line and `lty=2` will plot a dashed line.**   

```{r add-mean-med-plot}
# Expected solution [2 marks, where 1 point given for each line at the correct spot.]
ggplot(data = CS_data, aes(x = GDP_2006)) +
  geom_histogram(col = "white", fill = "sienna2") + 
  xlab("GDP in 2006") +
  theme_minimal() +
  geom_vline(aes(xintercept = 11790.67), lty=1) + 
  geom_vline(aes(xintercept = 3351.305), lty=2) 
```

```{r alternative solutions, eval = F, echo = F}
# Alternative solution ("base R" solution: students aren't expected to know this, 
# but some may write this solution based on previous R training)
ggplot(data = CS_data, aes(x = GDP_2006)) +
  geom_histogram(col = "white", fill = "sienna2") + 
  xlab("GDP in 2006") +
  theme_minimal() +
  geom_vline(aes(xintercept = GDP_summary$mean_GDP), lty=1) + 
  geom_vline(aes(xintercept = GDP_summary$median_GDP), lty=2) 

#Alternative solution ("tidyverse" solution, using dplyr's pull())
ggplot(data = CS_data, aes(x = GDP_2006)) +
  geom_histogram(col = "white", fill = "sienna2") + 
  xlab("GDP in 2006") +
  theme_minimal() +
  geom_vline(aes(xintercept = GDP_summary %>% pull(mean_GDP)), lty=1) + 
  geom_vline(aes(xintercept = GDP_summary %>% pull(median_GDP)), lty=2) 
```
\newpage

### Summarizing the distribution of cesarean delivery

Recall the distribution of cesarean delivery rates across countries:

```{r CS-histogram}
ggplot(data = CS_data, aes(x = CS_rate_100)) +
  geom_histogram(binwidth = 4, col = "white", fill = "sienna2") + 
  xlab("Cesarean delivery rate (%)") +
  theme_minimal()
```
\newpage

**7. [1 mark] Describe the shape of this distribution. Is it "skewed left",
"skewed right", "symmetric", or "bimodal"?**

Solution: [1 mark]: Bimodal. Students who say bimodal and right-skewed get the 
mark. Students who say only skewed right lose a mark.
\newpage

**There appears to be multiple peaks which sometimes points to there being another
variable that might explain the peaks. We can make a separate histogram for each
income group using the `facet_wrap()` function.**

**8. [1 mark, not auto-marked] Extend your ggplot code using the `facet_wrap()` statement to make a separate histogram for each level of the `Income_Group` variable:**

```{r CS-histogram-by-income}
#solution [1 mark for correct facet_wrap statement, facet_grid(. ~ Income_Group)
# or facet_grid(Income_Group ~ .) is also fine.]
ggplot(data = CS_data, aes(x = CS_rate_100)) +
  geom_histogram(binwidth = 6, col = "white", fill = "sienna2") + 
  xlab("Cesarean delivery rate (%)") +
  theme_minimal() + 
  facet_wrap(~ Income_Group)
```
\newpage

**9. [2 marks, not auto-marked] Based on this plot and the previous one, describe why the data had two peaks**

Solution: [2 marks, 1 point for discussing low/lower income, 1 point for 
discussing higher income countries] Many of the low and lower-middle income 
countries had CS rates < 10%, while the higher income counties have rates closer
to 20%.
\newpage

**10. [1 mark, not auto-marked] Why might lower income countries have lower rates of cesarean delivery?**

Solution: [1 mark for an answer that sounds reasonable, no marks if the answer 
doesn't make sense or is obviously incorrect] Lower income countries have 
reduced access to obstetrical care, especially surgical procedures, limiting the
number of women who can receive cesarean deliveries.
\newpage

**11. [2 marks] Modify the previously used `dplyr` `summarize()` code (from 
question 5) to calculate the mean and median of `CS_rate_100` using only one 
`summarize()` command. Assign this summary to the name `CS_summary` and then 
print the results by typing `CS_summary` so you can see the contents.**

```{r calc-CS-mean-median}
#solution [2 marks, 1 for each mean() and median() being calculated correctly]
CS_summary <- CS_data %>% summarize(mean_CS = mean(CS_rate_100),
                                    median_CS = median(CS_rate_100))

CS_summary
```
\newpage

**Measures of variation**

**12. [2 marks] Use ggplot2 to make a boxplot of the distribution of `CS_rate_100`**

```{r}
#solution [2 marks: set the correct y aes and used geom_boxplot to render a box plot. 
#Don't need to worry about the col, fill, or theme settings.]
ggplot(data = CS_data, aes(y = CS_rate_100)) +
  geom_boxplot(col = "black", fill = "sienna2") +
  theme_minimal()
```
\newpage

Recall that the box plot summarizes the distribution in five numbers: the minimum, 
the first quartile (with 25% of the data below it), the median, the third quartile 
(with 75% of the data below it), and the maximum. Each of these numbers has at 
least one corresponding R function:

| Number          | R function                                                   |
|-----------------|--------------------------------------------------------------|
| Minimum         | `min(variable)`                                              |
| First quartile  | `quantile(variable, probs = 0.25)`                           |
| Median          | `median(variable)` or `quantile(variable, probs = 0.5)`      |
| Third quartile  | `quantile(variable, probs = 0.75)`                           |
| Maximum         | `max(variable)`                                              |

**13. [2.5 marks] Use a combination of `dplyr`'s `summarize` function and the 
above functions to compute the five number summary of `CS_rate_100`. Assign the 
summary to the name `five_num_summary`:**

```{r five-number-summary}
#solution [2.5 marks: 0.5 mark for each of min(), Q1()... calculated correctly]
five_num_summary <- CS_data %>% summarize(
  min = min(CS_rate_100),
  Q1 = quantile(CS_rate_100, 0.25),
  median = median(CS_rate_100),
  Q3 = quantile(CS_rate_100, 0.75),
  max = max(CS_rate_100)
)

five_num_summary
```
\newpage

Double check that `geom_boxplot()` is making the box plot correctly. You can do 
this by adding horizontal lines to the plot at each number in your five number 
summary using `geom_hline()`. Because horizontal lines intercept the y-axis, 
`geom_hline()` requires the `yintercept` argument that you can set to each number 
in your summary. 

**14. [2 marks, not auto-marked] The code below includes one horizontal line at the minimum shown in blue. Add 
the rest of the lines:**

```{r add-horizontal-lines-to-boxplot}
#solution [2 marks, 0.5 mark for each 5.1, 15.6, 23.3, and 45.9 being added]
ggplot(data = CS_data, aes(y = CS_rate_100)) +
  geom_boxplot(col = "black", fill = "sienna2") +
  theme_minimal() + 
  geom_hline(aes(yintercept = 0.4), col = "blue") + 
  geom_hline(aes(yintercept = 5.1), col = "blue") +
  geom_hline(aes(yintercept = 15.6), col = "blue") +
  geom_hline(aes(yintercept = 23.3), col = "blue") +
  geom_hline(aes(yintercept = 45.9), col = "blue") 
```
\newpage

**15. [4 marks] Compile the following code which adds two points to the CS_data, 
makes a new dataset called `CS_data_plus_2`, and redraws the box plot. How did 
the box plot change? Perform a calculation to justify why it changed. What are 
the newly-added features on the plot called?**

```{r add-points, fig.width = 5}
out_data <- tibble::tribble(
  ~Country_Name, ~CS_rate_100,
  "Point 1", 90,
  "Point 2", 60
)

CS_data_plus_2 <- dplyr::full_join(CS_data, out_data)

ggplot(data = CS_data_plus_2, aes(y = CS_rate_100)) +
  geom_boxplot(col = "black", fill = "sienna2") 
```

Solution:
- [1 mark] There are two points above the top whisker on the revised box plot.
- [2 marks for the calculation] These points must be larger than Q3 + 1.5*IQR.

```{r calculations_new_IQR}
five_num_summary_new <- CS_data_plus_2 %>% summarize(
  min = min(CS_rate_100),
  Q1 = quantile(CS_rate_100, 0.25),
  median = median(CS_rate_100),
  Q3 = quantile(CS_rate_100, 0.75),
  max = max(CS_rate_100)
)
```
    - The IQR = Q3-Q1 =`r five_num_summary_new$Q3` - `r five_num_summary_new$Q1` = `r five_num_summary_new$Q3 - five_num_summary_new$Q1`. 
    - IQR times 1.5 = `r (five_num_summary_new$Q3 - five_num_summary_new$Q1) *1.5`
    - Q3 + `r 1.5*(five_num_summary_new$Q3 - five_num_summary_new$Q1)` = upper bound = `r 1.5*(five_num_summary_new$Q3 - five_num_summary_new$Q1)+five_num_summary_new$Q3`
    - Both 60 and 90 are larger than `r 1.5*(five_num_summary_new$Q3 - five_num_summary_new$Q1)+five_num_summary_new$Q3`, which is why they are suspected outliers.
- [1 mark] The points are called suspected outliers (or outliers is fine).

