---
title: "Assignment 6"
author: "Your name and student ID"
date: "Today's date"
output:
  pdf_document: default
---

* Due date: October 18, 5:00pm (make sure to provide enough time for Gradescope
submission to be uploaded if you include large image files).
* Late penalty: 50% late penalty if submitted within 24 hours of due date, no 
marks for assignments submitted thereafter.
* Submission process: Please submit a PDF of your assignment to Gradescope. You 
must tell Gradescope which questions are on which pages. If you can't see it 
properly on Gradescope, open the PDF in a PDF viewer at the same time so you 
can make the selections accurately. Not selecting, or selecting inaccurately, will result in point deductions.


Helpful hint:
- Knit your file early and often to minimize knitting errors! If you copy and 
paste code, you are bound to get an error that is hard to diagnose. Hand-writing 
code is the way to smooth knitting! We recommend knitting your file each time 
after you write a few sentences/add a new code chunk, so you can detect the 
source of knitting errors more easily.

--------------------------------------------------------------------------------

1. [9 points total] Oklahoma is not historically known for experiencing earthquakes. Up until 
2008, Oklahoma experienced a constant rate of about 1.5 perceptible earthquakes
per year on average.

a) [1 point] Assuming that earthquakes are random and independent, with a 
constant rate of 1.5 per year, the count of perceptible earthquakes per year in 
Oklahoma should have a Poisson distribution with mean 1.5. What is the standard 
deviation of the number of earthquakes per year?

Solution: SD = sqrt(variance) = sqrt(1.5) = 1.2247

b) [1 point] Making the same assumptions as in part (a), use one or two R functions to 
compute the probability of seeing less than two earthquakes per year. 

```{r}

#solution
#cumulative probability of seeing 1 or less 
ppois(q = 1, lambda = 1.5) 

#sum of the probability of seeing exactly 1 and the probability of seeing 0
dpois(x = 1, lambda = 1.5) + dpois(x = 0, lambda = 1.5)
```

c) [2 points] Do the same calculation as above, this time using only a hand 
calculator. Show your work and round your final percentage to two decimal places.

Solution:
$P(X=k) = \frac{e^{-\mu}\mu^k}{k!}$

$P(X=0) = \frac{e^{-\mu}\mu^0}{0!}=e^{-1.5}=0.2231302$

$P(X=1) = \frac{e^{-1.5}1.5^1}{1!}=0.3346952$

Thus: $P(X<2) = P(X=0) + P(X=1) = 0.2231302 + 0.3346952 = 0.5578254 = 55.78\%$

d) [1 point] In 2013, Oklahoma experienced 109 perceptible earthquakes (an 
average of two per week). Assuming the same model as above, write an equation to
show how the chance of experience 109 earthquakes or more can be written as a 
function of the probability at or below some $k$. 

(Note: You can write these equations using pen and paper and upload the image if
you'd like. You can also write the equations using plain text (i.e., P(X >= k)).
If you would like to use math equations that render when you knit the pdf 
(i.e., $P(X>=k)$) you need to be **very careful** with your symbols. For example, to
get the symbol for "greater than or equal to" you cannot copy and paste it into 
R from the slides or another document. This will cause errors! Instead you need to
write $P(X \geq k)$. Again, use any of these three methods (hand-written, plain
text in R, or "math equations between dollar signs", and you will get points 
so long as it is human-readable.) 

Solution: $P(X \geq 109) = 1 - P(X \leq 108)$

e) [2 points] Use R, calculate the probability of observing 109 perceptible 
earthquakes or more. Write a sentence describing the chance of seeing such an 
event assuming the specified Poisson distribution (i.e., is it rare or common?)

```{r}
#write your code here. 

#solution a (At or above k=109 is equal 1 - at or below k = 108.):
1 - ppois(q = 108, lambda = 1.5, lower.tail = T)

#solution b (Use the upper tail probability at or above 109):
ppois(q = 109, lambda = 1.5, lower.tail = F)
```

f) [2 points] Based on your answer in question (e), would you conclude that the 
mean number of perceptible earthquakes has increased? Why or why not? Would 
knowing that the number of perceptible earthquakes was 585 in 2014 support your 
conclusion?

Solution: [1 point for correct conclusion. 1 point for explanation] Yes the mean number of perceptible earthquakes has increased. The probability of observing such a high number of earthquakes is essentially 0 when the true mean is 1.5 earthquakes per year. Yes observing 585 earthquakes in 2014 supports my conclusions that the true mean is increasing.

\newpage

2. [5 points total] To track epidemics, the Center for Disease Control and Prevention requires 
physicians to report all cases of important transmissible diseases. In 2014,
a total of 350,062 cases of gonorrhea were officially reported, 53% of which 
were individuals in their 20s. Assume this 53% stays the same every year.
Researchers plan to take a simple random sample of 400 diagnosed cases of 
gonorrhea to study the risk factors associated with the disease. Call $\hat{p}$
the proportion of cases in the sample corresponding to individuals in their 20s. 

a) [2 points] What are the mean and standard deviation of the sampling 
distribution of $\hat{p}$ in random samples of size 400?

Solution:
mean = 0.53, since the sample mean is an unbiased estimator of $p$
standard deviation = sqrt(p(1-p)/n) = sqrt(0.53(1-0.53)/n) = 0.02495496

The mean of the sampling distribution is 0.53 and the standard deviation is 
approximately 0.025 when the sample is size 400.

b) [3 points] Describe the conditions required for the sampling distribution of $\hat{p}$ 
to be Normally distributed. Use the numbers provided in the question to check if
the conditions are likely met. 

- The population is expected to be at least 20 times larger than the sample. 
Using the 2014 data, the population of >350k cases is much much larger that a 
sample of size 400
- $400\times 0.53 = 212$, and $400 \times (1-0.53) = 188$ are both greater than 10, 
implying that n is large enough and p is not too rare or too common.
- Yes the conditions are met for the distribution of $\hat{p}$ to be Normally
distributed.

\newpage

3. [4 points total] Read [this short article](https://www.nytimes.com/2016/10/06/upshot/when-you-hear-the-margin-of-error-is-plus-or-minus-3-percent-think-7-instead.html)
in the New York Times Upshot from 2016:

a) [2 points] Explain sampling variation, in the context of this article. Does the
3 percentage point margin of error account for sampling variation?

Solution: 

Sampling error occurs here because the survey of voters is based on a sample of
the total voting population. 

Yes, the margin of error accounts for sampling variation.

b) [1 point] The authors provides several reasons why the true margin of error is 
larger than three percent. Describe one of the primary reasons provided in 1-2 
sentences.

Any of:

- "Frame error": mismatch between people who were polled vs. true target population.
- Nonresponse error: likelihood of responding is systematically related to how 
one would have answered the survey.
- "Analysis error": pollsters are performing the analysis wrong.

Note: students may not use these exact terms, but we're looking for them to describe
one of these three errors. 

c) [1 point] Based on the information in article, if we're doing a study in public
health, choose the answer that is most correct:

    (a) The confidence interval accounts for random error only. If a study suffers
    from other sources of bias (i.e., confounding, or mismeasurement) the CI will
    not account for this limitation.
    (b) Increasing the sample size will reduce the chance of other sources of bias
    (i.e., confounding, or mismeasurement), which is why a larger sample is better.
    (c) both (a) and (b)
    d) neither (a) or (b)

Solution: (a) is correct. 

Note: (b) is incorrect because increasing sample size does 
nothing to remove systematic biases like confounding or mismeasurement. 

\newpage

4. [11 points total] Deer mice are small rodents native in North America. Their adult body lengths
(excluding tail) are known to vary approximately Normally, with mean $\mu=86$ mm
and standard deviation $\sigma=8$ mm. It is suspected that depending on their 
environment, deer mice may adapt and deviate from these usual lengths. A random 
sample of $n=14$ deer mice in a rich forest habitat gives an average body length 
of $\bar{x}=91.1$ mm. Assume that the standard deviation $\sigma$ of all deer 
mice in this area is 8 mm. 

a) [2 points] Calculate a 99% confidence interval based on this information (you can use R 
as a calculator to perform the calculation, or use a hand calculator. Either way,
show your work). Interpret the confidence interval. Round your final values to 
two decimal places.

```{r CI99}
#write your code here (none needed if you use hand calculator)

#solution
known.sigma <- 8
critical.value <- 2.576

91.1 - critical.value*(8/sqrt(14)) #lower bound
91.1 + critical.value*(8/sqrt(14)) #upper bound
```

Our 99% CI for this population of deer mice lengths is 85.59mm to 96.61mm. This means
that if we were to take 100 samples using this same method, 99 of them would contain
the true value $\mu$ in the underlying population and 1 would not. 

b) [2 points] Suppose deer mice researchers thought your CI was too wide to be 
useful. Given that you cannot change the standard deviation, what two things 
could you do to provide a narrower confidence interval? 

Solution:
- Reduce the level of confidence from 99% to 95% or to 90% even.
- Increase the sample size

c) [3 points] You decide to create a 95% confidence interval, rather than a 99% 
confidence interval. Perform this calculation below. If performing a hand calculation, 
show your work. Based on this 95% CI, is there evidence against the hypothesis 
$H_0$ that these mice have a significantly different mean length compared to the 
population described in the first part of the question? 
Without performing a calculation, what amounts do you know the p-value to be 
bounded between for a two-sided hypothesis test of $H_0$? 

Hint: 
Use information from this question and from part (a).

Solution 

```{r CI95}
#write your code here (none needed if you use hand calculator)

#solution
known.sigma <- 8
critical.value <- 1.96

91.1 - critical.value*(8/sqrt(14)) #upper bound
91.1 + critical.value*(8/sqrt(14)) #upper bound
```

The 95% confidence interval is from 86.91mm to 95.29mm. Thus, there is evidence
against $H_0: \mu = 86$, because 86mm is not contained within this 95% confidence
level. We know that the p-value is greater then 0.01 but less than 0.05 because 
86mm is outside of the 95% confidence interval but inside the 99% confidence 
interval.

d) [4 points] Perform a z-test with the two-sided alternative hypothesis that 
$H_a: \mu \neq 86$ (and null hypothesis $H_0: \mu = 86$). Be sure to write out 
the null and alternative hypotheses, the z test statistic, the p-value and an
interpretation of the p-value.

Solution: 

$z = \frac{\bar{x}-\mu}{\sigma/\sqrt{n}} = \frac{91.1-86}{8/\sqrt{14}} = 2.338536$

```{r ztest}
2*pnorm(2.338536, mean = 0, sd = 1, lower.tail = F)
```

p-value = 0.01935946 = 1.94%

Thus, there is a 1.94% chance of seeing a results this or more extreme under the
null hypothesis of no difference in the means. 


