---
title: "Assignment 7"
author: "Your name and student ID"
date: "Today's date"
output:
  pdf_document: default
---

```{r, warning=F, message=F, echo=F}
library(dplyr)
library(ggplot2)
```

* Due date: November 1, 5:00 pm (make sure to provide enough time for Gradescope
submission to be uploaded if you include large image files).
* Late penalty: 50% late penalty if submitted within 24 hours of due date, no 
marks for assignments submitted thereafter.
* Submission process: Please submit a PDF of your assignment to Gradescope. You 
must tell Gradescope which questions are on which pages. If you can't see it 
properly on Gradescope, open the PDF in a PDF viewer at the same time so you 
can make the selections accurately, others points will be deducted since this 
makes grading much less efficient.

Helpful hint:

- Knit your file early and often to minimize knitting errors! If you copy and 
paste code, you are bound to get an error that is hard to diagnose. Hand-writing 
code is the way to smooth knitting! We recommend knitting your file each time 
after you write a few sentences/add a new code chunk, so you can detect the 
source of knitting errors more easily.
 
--------------------------------------------------------------------------------

\newpage

**Question 1 [12 points total].** 

In two wards for elderly patients in a local hospital, the following 
levels of hemoglobin (grams per liter) were found for a simple random sample of
patients from each ward.:

Ward A: 
```{r}
ward_a <- c(12.2, 11.1, 14.0, 11.3, 10.8, 12.5, 12.2, 11.9, 13.6, 12.7, 13.4, 13.7)
```

Ward B: 
```{r}
ward_b <- c(11.9, 10.7, 12.3, 13.9, 11.1, 11.2, 13.3, 11.4, 12.0, 11.1)
```

1.a) [2 points] Make two box plots to compare the hemoglobin values for Ward A and Ward B, and display it.
Overlay the boxplots with their raw data. Comment on the similarities/differences
portrayed by the plots, keeping in mind that the sample size is relatively small
for these two wards.

```{r make-data-frame}
hemoglobin <- data.frame(hemo_level = c(ward_a, ward_b), 
                         ward = c(rep("Ward A", 12), rep("Ward B", 10)))
```

```{r make-box-plot}
# Your code here.
```

Write your answer here.

\newpage

1.b) [2 points] What two assumptions do you need to make to use any of the t-procedures? 
Because each ward has a rather small sample size (n < 12 for both), what two 
characteristics of the data would you need to check for to ensure that the 
t procedures can be applied?

Write your answer here.

\newpage

1.c) [4 points] Using only `dplyr` and `*t` functions, create a 95% confidence interval 
for the mean difference between Ward A and Ward B. You can do this by using `dplyr`
to calculate the inputs required to calculate the 95% CI, and then plugging these
values in on a separate line of code (or using your calculator). Use a degrees of
freedom of 19.515 (You don't need to calculate the degrees of freedom, you can 
use this value directly). Show your work and the result, and interpret the mean difference and its 95% CI.

```{r}
# Your code here.
```

Write your answer here.

\newpage

1.d) [4 points] Perform a two-sided t-test for the difference between the two samples, where
the null hypothesis is that the underlying means are the same. Start by 
writing down the null and alternate hypotheses, then calculate the test statistic
(showing your work) and p-value. Continue to assume that the degrees of freedom
is 19.515. Verify the p-value by running the t-test using R's built in function.
Show the output from that test. Hint: to perform the t-test using R's built in 
function, you need to pass the function an x and y argument, where x includes 
that values for Ward A and Y includes the values for Ward B. `dplyr`'s `filter()`
and `pull()` functions will be your friends.

(Hypothesis) Write your answer here.

```{r test statistics and p-value}
# Your code here.
```

(Conclusion) Write your answer here.

```{r t-test}
# Your code here.
```

\newpage

**Question 2 [7 points total]**

The time to perform open heart surgery is normally distributed. Sixteen 
patients (chosen as a simple random sample from a hospital) underwent open heart 
surgery that took the following lengths of time (in minutes).

```{r}
op_time <- c(247.8648, 258.4343, 315.6787, 268.0563, 269.9372, 320.6821, 
             280.5493, 225.3180, 243.8207, 251.5388, 304.9706, 277.3140, 
             278.6247, 269.3418, 248.0131, 322.9812)
surg_data <- data.frame(op_time)
```

a) [1 point] You wish to know if the mean operating time of open heart surgeries at this 
hospital exceeds four hours. Set up appropriate hypotheses for investigating
this issue.

Write your answer here.

\newpage

b) [3 points] Test the hypotheses you formulated in part (a). Report the p-value. What are 
your conclusions? (Do not use the `t.test` function for this question)

```{r}
# Your code here.
```

Write your answer here.

\newpage

c) [3 points] Construct a 95% CI on the mean operating time.

Write your answer here.


\newpage

3. [1 point] Suppose you were testing the hypotheses $H_0: \mu_d = 0$ and $H_a: \mu_d \neq 0$
in a paired design and obtain a p-value of 0.21. Which one of the following could
be a possible 95% confidence interval for $\mu_d$?

a. -2.30 to -0.70
b. -1.20 to 0.90
c. 1.50 to 3.80
d. 4.50 to 6.90

Write your answer here.

\newpage

4. [1 point] Suppose you were testing the hypotheses $H_0: \mu_d = 0$ and $H_a: \mu_d \neq 0$
in a paired design and obtain a p-value of 0.02. Also suppose you computed confidence
intervals for $\mu_d$. Based on the p-value which of the following are true?

a. Both a 95% CI and a 99% CI will contain 0.
b. A 95% CI will contain 0, but a 99% CI will not.
c. A 95% CI will not contain 0, but a 99% CI will.
d. Neither a 95% CI nor a 99% CI interval will contain 0. 

Write your answer here.
