---
title: "Assignment 8"
author: "Your name and student ID"
date: "Today's date"
output:
  pdf_document: default
---

* Due date: November 8, 5:00pm (make sure to provide enough time for Gradescope
submission to be uploaded if you include large image files).
* Late penalty: 50% late penalty if submitted within 24 hours of due date, no 
marks for assignments submitted thereafter.
* Submission process: Please submit a PDF of your assignment to Gradescope. You 
must tell Gradescope which questions are on which pages. If you can't see it 
properly on Gradescope, open the PDF in a PDF viewer at the same time so you 
can make the selections accurately, otherwise points will be deducted since this 
makes grading much less efficient.

Helpful hint:

- Knit your file early and often to minimize knitting errors! If you copy and 
paste code, you are bound to get an error that is hard to diagnose. Hand-writing 
code is the way to smooth knitting! We recommend knitting your file each time 
after you write a few sentences/add a new code chunk, so you can detect the 
source of knitting errors more easily. This will save you and the GSIs from frustration!

- If your code runs off the page of the knitted PDF then you will LOSE POINTS! To
avoid this, have a look at your knitted PDF and ensure all the code fits in the 
file. When it doesn't, go back to your .Rmd file and add spaces (new lines) using
the return or enter key so that the code runs onto the next line.
 
 \newpage

Question 1 [3 points total]. You would like to conduct a survey of highschool students to determine
the proportion who are current e-cigarettes users. Before you conduct your 
survey, you need to determine how large of a sample size you need. Suppose that you would 
like the **width** of the 95% confidence interval to be 5 percentage points.

1.a) [1 point] Determine the most conservative sample size you would require. Recall that
to do this, you need to use a $p^*$ of 0.5.

Write your answer here.

\newpage

1.b) [1 point] You've seen a recent publication ([link](http://annals.org/aim/fullarticle/2698112/prevalence-distribution-e-cigarette-use-among-u-s-adults-behavioral))
from the *Annals of Internal Medicine* that estimated that 9.2% of individuals 
aged 18 to 24 years old are current e-cigarette users. Update your sample size
estimate under the assumption that the true underlying parameter $p=0.092$. 

Write your answer here.

\newpage

1.c) [1 point] The recent publication referenced in the previous question only looked at 
adults (aged 18+), but observed that the rate of current use was inversely 
related to age among the population they surveyed. Because of this finding
would you suppose that the sample size estimated in part (b) is too low or too
high?

Write your answer here.

\newpage

Question 2 [12 points total]. Exclusive breastfeeding during the first six months 
of life is recommended for optimal infant growth and development. Suppose that 
you conducted a survey of randomly chosen women from California and found that 
775 out of 5615 new mothers exclusively breastfed their infants.

2.a) [4 points]. Perform all four of the methods discussed in lecture and during
lab to create a 95% confidence interval for the proportion of infants 
exclusively breast fed. Remember that two of these methods can be performed "by hand"
and two using R functions (though you can use R as a calculate for all four 
methods if you wish).

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

```{r}
#your code here.
```

Summarize your findings: 

- 95% CI using _____ method: 
- 95% CI using _____ method: 
- 95% CI using _____ method: 
- 95% CI using _____ method: 

\newpage

2.b) [3 points] Create a plot comparing the confidence intervals. Do the
methods produce confidence intervals that are basically the same or very different?
Why?

Write your answer here.

```{r}
#your code here.
```

\newpage

2.c) [1 point] Suppose that in 2010, the rate of exclusive breastfeeding in 
California was known to be 18.6%. Based on the 95% CIs calculated in 2.b) is there
evidence against the null hypothesis that the underlying rate is equal to 18.6% in 
favor of the alternative that the rate is different from 18.6%?

Write your answer here.

\newpage

2.d) [4 points] To confirm your answer to part c), perform a two-sided hypothesis
test and interpret the p-value. 

State the null and alternative hypotheses:

Write your answer here.

Calculate the test statistic:

Write your answer here. You may or may not want to insert a code chunk.

Calculate the p-value:

```{r}
#your code here.
```

Interpret the p-value:

Write your answer here.

\newpage

Question 3 [2 points total]. The quadrivalent HPV vaccine was introduced to
Canada in 2007, and was given to girls in Ontario, Canada who were enrolled in 
grade 8 (13-14 year olds). Before 2007, no girls recieved the vaccine, while in 
the 4 years after it was introduced nearly 40% of girls recieved the vaccine 
each year. One concern that some people had was that the vaccine itself would 
increase promiscuity if the girls felt a false sense of protection, which could 
thereby increase the prevalence of other sexually transmitted
infections (STIs) among vaccinated girls. A paper ([link](https://www.ncbi.nlm.nih.gov/pubmed/25487660))
examines this question using an advanced method called the "regression discontinuity"
design which harnesses the abrupt change in vaccination status across cohorts
of girls to estimate the causal effect of vaccination against HPV on the 
occurrence of other STIs.

Read only the abstract of the paper, and don't worry about the 
details because these are advanced methods. Note that the term "RD" is the difference
in risk of STIs between girls exposed and unexposed to HPV vaccination. We can
therefore think of this risk difference as the difference between two proportions.

3.a) Interpret this result from the abstract: We identified 15 441 (5.9%) cases 
of pregnancy and sexually transmitted infection and found no evidence that 
vaccination increased the risk of this composite outcome: RD per 1000 
girls -0.61 (95% confidence interval [CI] -10.71 to 9.49)

[1 point] In particular, what does -0.61 estimate?

Write your answer here.

\newpage

3.b) [1 point] The 95% confidence interval includes 0. What can you conclude about the
p-value for a two-sided test of the difference between vaccinated and unvaccinated
girls and their risk of sexually transmitted diseases?

Write your answer here.

\newpage

Question 4 [7 points total]. An allergy to peanuts is increasingly comment in Western
countries. A randomized controlled trial enrolled infants with a diagnosed peanut
sensitivity. Infants were randomized to either avoid peanuts or consume them
regularly until they reached age 5. At the end of the study 18 out of the 51 
randomized to avoidance were allergic to peanuts compared to 5 out of the 47 
randomized to consuming them regularly. 

4.a) [1 point] Estimate the difference between the two proportions.

Write your answer here. You can insert an R chunk to perform calculations if you
wish.

\newpage

4.b) [1 point] Use the plus four method to find a 99% confidence interval for the difference
between the two groups. 

Write your answer here. You can insert an R chunk to perform calculations if you
wish.

\newpage

4.c) [1 point] Why would it have been inappropriate to use the large sample method to create
a 99% CI?

Write your answer here.

\newpage

4.d) [4 points] Perform a two-sided hypothesis test for the difference between the groups. 
Start by stating the null and alternative hypotheses, then calculate the test
statistic, the p-value, and conclude with your interpretation of the p-value.

Write your answer here.

\newpage

5. [1 point] Suppose you were testing the hypotheses $H_0: \mu_d = 0$ and $H_a: \mu_d \neq 0$
in a paired design and obtain a p-value of 0.21. Which one of the following could
be a possible 95% confidence interval for $\mu_d$?

a. -2.30 to -0.70
b. -1.20 to 0.90
c. 1.50 to 3.80
d. 4.50 to 6.90

Write your answer here.

\newpage

6. [1 point] Suppose you were testing the hypotheses $H_0: \mu_d = 0$ and $H_a: \mu_d \neq 0$
in a paired design and obtain a p-value of 0.02. Also suppose you computed confidence
intervals for $\mu_d$. Based on the p-value which of the following are true?

a. Both a 95% CI and a 99% CI will contain 0.
b. A 95% CI will contain 0, but a 99% CI will not.
c. A 95% CI will not contain 0, but a 99% CI will.
d. Neither a 95% CI nor a 99% CI interval will contain 0. 

Write your answer here.
