---
title: "Assignment 9"
author: "Your name and student ID"
date: "Today's date"
output:
  pdf_document: default
---

* Due date: November 22, 5:00pm (make sure to provide enough time for Gradescope
submission to be uploaded if you include large image files).
* Late penalty: 50% late penalty if submitted within 24 hours of due date, no 
marks for assignments submitted thereafter.
* Please do not change the format of the assignment. This makes grading less efficient.
* Submission process: Please submit a PDF of your assignment to Gradescope. You 
must tell Gradescope which questions are on which pages. If you can't see it 
properly on Gradescope, open the PDF in a PDF viewer at the same time so you 
can make the selections accurately. Not selecting, or selecting inaccurately
will result in points being deducted since this makes grading much less efficient.

Helpful hint:

- Knit your file early and often to minimize knitting errors! If you copy and 
paste code, you are bound to get an error that is hard to diagnose. Hand-writing 
code is the way to smooth knitting! We recommend knitting your file each time 
after you write a few sentences/add a new code chunk, so you can detect the 
source of knitting errors more easily.
 
\newpage
--------------------------------------------------------------------------------

Question 1. [7 marks total] Parental leave is often compensated to some degree, but the amount
of compensation varies greatly. You read a research article that stated, 
"across people of all incomes, 47% of leave-takers received full pay during 
their leave, 16% received partial pay, and 37% received no pay." 

After reading this, you wonder what the distribution of parental leave payment 
pay is for low income households. Suppose you conduct a survey of leave-takers within households 
earning less than $30,000 per year. You surveyed 225 people (selected in a 
random sample) and found that 51 received full pay, 33 received partial pay, and
141 received no pay.

1.a) [1 mark] You would like to investigate whether the distribution of pay for 
households earning < $30,000 is different from that of all income levels. Does 
this correspond to a chi-square test of independence or a chi-square test for 
goodness of fit?

Your answer here.


\newpage

1.b) [2 marks] What are the expected counts of leave-takers among households with
incomes < $30,000. State the null hypothesis under which these expected counts
were computed.

Your answer here.


\newpage

1.c) [2 marks] Compute the chi-square statistic and comment on the cell (i.e. term in summation) that contributes
the most to the statistic. 

Your answer here.


\newpage

1.d) [2 marks] Compute the p-value for your test statistic and conclude whether 
you believe there is evidence against the null hypothesis in favor of the alternative
hypothesis.

Your answer here.


\newpage


Question 2. [9 marks total] Human papillomavirus (HPV) is a very common STI that most sexually
active persons will encounter during their lifetimes. While many people clear 
the virus, certain strands can lead to adverse health outcomes such as 
genital warts and cervical cancer. 

Suppose that you selected a random sample from a population and collected
these data on age and HPV status for the sample:

|Age Group    | HPV +        | HPV -          | Row total   |
|:-----------:|:------------:|:--------------:|:-----------:|
|14-19        | 160          | 492            | 652 (33.9%) |
|20-24        | 85           | 104            | 189 (9.8%)  |
|25-29        | 48           | 126            | 174 (9.1%)  |
|30-39        | 90           | 238            | 328 (17.1%) |
|40-49        | 82           | 242            | 324 (16.9%) |
|50-59        | 50           | 204            | 254 (13.2%) |
| Col total   | 515 (26.8%)  | 1406 (73.2%)   | 1921        |

2.a) [1 mark] Which variable is explanatory and which is response?

Your answer here.


\newpage

2.b) [2 marks] Formulate null and alternative hypotheses using these data to test
whether there is a relationship between age group and HPV status. State these 
hypotheses using the language or notation of conditional distributions.

Your answer here.


\newpage

2.c) [1 mark] Run the code below to examine the conditional distribution of age by HPV
status. Based on this plot, which age group will contribute the most to the 
chi-square statistic? (That is, can you tell based on this plot when the observed
count will differ most from the expected count under the null hypothesis of no
relationship between age group and HPV status?) 

Your answer here.


```{r create-hpv.data, echo=F, out.width="80%", fig.align=T, message=F}
age_group <- rep(c("14-19", "20-24", "25-29", "30-39", "40-49", "50-59"), 2)
hpv_test <- c(rep("positive", 6), rep("negative", 6))
number_of_women <- c(160, 85, 48, 90, 82, 50, 
                     492, 104, 126, 238, 242, 204)

hpv_data <- data.frame(age_group, hpv_test, number_of_women)

library(tidyverse)
hpv_data <- hpv_data %>% group_by(hpv_test) %>% 
  mutate(sum_women = sum(number_of_women),
         prop_women_age = number_of_women/sum_women)

ggplot(hpv_data, aes(x = age_group, y = prop_women_age)) + 
  geom_bar(aes(fill = hpv_test), stat = "identity", position = "dodge") +
  theme_minimal(base_size = 15) +
  labs(y = "Proportion", x = "Age", 
       main = "Conditional distribution of age by HPV status")
```



\newpage

2.d) [2 marks] Fill out the table of expected counts under the null hypothesis of no 
association between age group and HPV status. You don't need to show your 
work, but make sure you can calculate the expected counts by hand, using a 
calculator.

Expected counts:

|Age Group    | HPV +        | HPV -        | 
|:-----------:|:------------:|:------------:|
|14-19        | ?            | ?            | 
|20-24        | ?            | ?            | 
|25-29        | ?            | ?            | 
|30-39        | ?            | ?            | 
|40-49        | ?            | ?            |
|50-59        | ?            | ?            |



\newpage

2.e) [3 marks] Calculate the test statistic and p-value and assess whether there is 
evidence against the null in favor of the alternative.

Your answer here.

```{r}
# your code here
```



\newpage


Question 3. [7 marks] Fill in the blanks. Please let the asterisks so your answers
are bolded in the rendered file.

The bootstrap method is used to compute **_______________**, while the permutation
test is used to conduct **_______________**.

Bootstrapping involves taking repeated simple random samples **_______________** replacement
from the original sample of the **_______________** size as the original 
sample. For each bootstrap, the statistic of interest is calculated (say the median).
These bootstrapped statistics are then plotted on a **_______________** to visualize their distribution and the 
**_______________** and **_______________** quantiles are computed to calculate a
95% confidence interval.


