---
title: "Assignment 10: Regression of voting patterns across 3141 US counties"
author: "Your name and student ID"
date: "Today's date"
output:
  pdf_document: default
---

* Due date: Dec 2, 5:00pm (make sure to provide enough time for Gradescope
submission to be uploaded if you include large image files).
* Please do not change the format of the assignment. This makes grading less efficient.
* Submission process: Please submit a PDF of your assignment to Gradescope. You 
must tell Gradescope which questions are on which pages. If you can't see it 
properly on Gradescope, open the PDF in a PDF viewer at the same time so you 
can make the selections accurately. Not selecting, or selecting inaccurately
will result in points being deducted since this makes grading much less efficient.

Helpful hint:

- Knit your file early and often to minimize knitting errors! If you copy and 
paste code, you are bound to get an error that is hard to diagnose. Hand-writing 
code is the way to smooth knitting! We recommend knitting your file each time 
after you write a few sentences/add a new code chunk, so you can detect the 
source of knitting errors more easily.
 
\newpage
--------------------------------------------------------------------------------

### Voting during the 1992 election

In the spirit of the upcoming 2020 presidential election, I thought it would be 
interesting to consider some historical data on voting patterns across US counties.

This code loads in the data frame `counties`:

```{r}
load("../Data/A10_counties.sav")
```

These data are from the 1992 election and look at the percent of votes cast 
(for each county) for the `democrat` (Bill Clinton), `republican` (George Bush),
and independent presidential nominees (Ross `Perot`). 

Ideally, if you were interested in voting patterns, you might look at the 
relationship between individual characteristics and whether each individual 
voted Democrat or Republican. However, data like that is often hard to come by.
The `counties` data provide data on 3141 counties. Use `View()` to examine these
data briefly and read the labels corresponding to the variables. Note that Alaska
is not included and that two other counties with populations = 0 have also been 
excluded.

As discussed in class we have the entire population (not just a sample), so 
strictly speaking we don't need to perform statistical inference. However, we
might pretend this is a sample so that we can apply the techniques of inference 
and gain competence creating and interpeting a linear model.

\newpage
Question 1 [2 marks]. Looking only at California, plot the relationship between 
the % of votes cast for the Democratic candidate (`democrat`) 
and the population density of the county (`pop.density`). 

```{r}
# Your code here.
```

\newpage
Question 2 [1 mark]. 
The above plot you made does not look very good for question 1. 
The distribution of population density is skewed right, 
with a few counties having much higher densities than the majority of counties.
To see which counties these are, we will use `geom_text_repel` from the 
library `ggrepel`. 
The template for using this function is: 
`geom_text_repel(aes(label = your_labelling_var))`. 
You will want to set the labeling variable to be the variable in the dataset
containing the county names.

```{r}
library(ggrepel)

# Your code goes here.
```

The current issue with these data is that San Francisco 
(as you can now hopefully point out) 
has a much higher population density than other counties, 
and that generally there is a large right skew in the distribution 
of the population density variable. 

If we tried and fit a linear model to these data, it would not fit well-- because
the relationship between population density and the response variable is not 
linear. However, this is the perfect situation to try transforming the x variable.

\newpage
Question 3. [2 marks] Try adding a log-transformed version of population density
to the data frame and remake your plot using this new variable. Keep the 
population labels. Also add a smoothed fitted line:

```{r}
# Your code goes here.
```

\newpage
Question 4 [4 marks]. Describe the relationship between the (logged) population density 
and the response variable in terms of the shape, direction, strength, and outliers.
These are concepts from Chapter 3. Calculate the correlation to
comment on one of these aspects.

<Your description here.>

\newpage
Question 5 [4 marks] Run a linear model regression of the % votes cast for the democratic
candidate as a function of the population density. Make sure you get the order
of variables right in the `lm()` function! Use a `broom()` command to show the 
slope and intercept estimates. Interpret the association between the logged population 
density and the response variable. (You can `View()` the data frame to make sure
you are getting your units right by checking the descriptions in the labels for 
each variable). Report and interpret the r-squared for the model.

```{r}
# Your code here.
```

<Your solution goes here.>

\newpage
Question 6 [4 marks]. Using the code learned in class, that was also shown in 
last week's lab, make the four plots to examine the assumptions. 

```{r}
# Your code goes here.
```

\newpage
Question 7. [4 marks] Comment on each of the plots and conclude about which 
assumptions appear violated vs. not violated. Don't forget to comment on the 
one assumption that cannot be investigated using plots.

<Your comments here>.

\newpage
-------------------------------------------------------------------------------

### Part 2: Choose your own adventure [10 marks for completion]

Using the `counties` data frame pick a state (California if you still want to, or
another one) and investigate the relationship between voting patterns and an
explanatory variable of your choice. You can keep using `democrat` as the response 
variable if you want to further your investigation of predictors of this 
response, or you can change it to `republican`, or `Perot` (the independent 
candidate). Set your explanatory variable to one of:

- `college`
- `crime`
- `white`
- `black`
- `farm`
- `age6574` or `age75`, or a `mutate`d variable based on age combining across these two age variables.
- `income`

Following questions above, your task:

- Make a scatter plot of the relationship, with labels for the counties and a 
smoothed fitted line added to the plot. Link the size of the points to population
density.
- Comment on the relationship in terms of form, direction, strength, and outliers.
- Even if the relationship is clearly not linear, make the residual plots anyway
and comment on whether you can catch the non-linearity or a violation of one of 
the other model assumptions using the plots.
- 10 marks for the full question with marks deducted for insufficient examination in any of these steps.
