---
title: "Lab03: Simulation (optional)"
author: "Your Name and Student ID"
date: "Today's Date"
output: pdf_document
---

# Part II: Simulation as tool for understanding statistics

This section contains slightly advanced material for those interested in 
learning about using R for statistical simulations. Therefore, part II is
optinal and not required to be submitted for grading.

One very useful feature of R is the ability to create simulations, both as
research and educational tools. More specifically, one can define a set of 
functions to generate "realistic" data, and then observe the behavior of 
algorithms used to estimate statistics based on the simulated data. The benefit 
of simulating data vs. using actual data, is that you now know the true values 
the statistics should take because you defined them yourself. (The downside of 
simulations, is they usually are an over-simplistic representation of the real 
process that generated the data). Simulations are used for a wide variety of
reasons. For example, you could use a simulation to investigate whether the spread
of an infectious disease that has occurred in the real world is aligned with a
model that you have proposed. 

This part of the lab begins to introduce you to the steps followed to simulate
data using R. Later in the course, we will use simulations to discuss estimation, 
inference and sampling distributions. 

For now, we generate data from a linear model $Y = a + bX + e$, where $e$ is a 
random error (otherwise, all points would lie exactly on the line).

```{r define-functions}
# runif() makes n copies of random number between 0 and 2
# Run this line and then take a look at X by printing it to the screen
X <- runif(n = 100, min = 0, max = 2)
X # Notice that there are 100 values between 0 and 2. 

# Make a random error that is normally distributed 
# We will learn more about the normal distribution in Chapter 11.
e <- rnorm(n = 100, mean = 0, sd = 0.25)

# Make Y, the reponse variable, a function of X and e
sim_intercept <- 0 # we arbitrarily set the intercept to 0 to begin. 
sim_slope <- 0.5 # we arbitrarily set the slope to 0.5 to begin. 
Y <- sim_intercept + sim_slope*X + e
dat_sim <- data.frame(Y, X)
```

**11. Make a scatter plot of Y versus X, variables in the data set `dat_sim`.**
```{r plot-the-data}
# use this code chunk to write your code.
```

**12. Run a linear regression of Y on X.**
```{r run-regression}
# use this code chunk to write your code.
```

**13. Investigate what happens if you specify `sim_slope` to equal -1 rather than
0.5. You can do this by overwriting the contents of `sim_slope` so that it equals 
-1 and then re-creating `Y` and `dat_sim` by copying and pasting the lines of 
code that define `Y` and `dat_sim`, making sure to run the new lines of code.
Also re_create the plot and re-run the linear regression. What changes?**

```{r rinse-and-repeat}
# use this code chunk to write your code.
```

Replace this text with your answer.
