---
title: "Lab 3: Regression"
author: "Your Name and Student ID"
date: "Today's Date"
output: pdf_document
---

### Instructions

This lab uses the same data set as Lab 1 and Assignment 1. In those exercises, you
investigated the distribution of global cesarean delivery rates and country's 
GDPs. In part I of this lab, we will look at the relationship between two 
variables using skills we learned from Chapters 3 and 4. (Part II on
simulations is optional.)

Start by loading the required libraries, reading in the data and adding on a 
variable:

```{r initial-setup}
library(dplyr)
library(ggplot2)
library(readr)
library(broom)

CS_data <- read_csv("../../Data/Cesarean.csv")

# The following code re-orders the variable Income_Group in the specified order.
# Note that it *does not* change the order of the data frame (like arrange() does)
# Rather, it specifies the order the data will be plotted. 
# This will make more sense when we plot the data using Income_Group, and then
# again using Income_Group_order
CS_data$Income_Group <- forcats::fct_relevel(CS_data$Income_Group, 
                                            "Low income", "Lower middle income", 
                                            "Upper middle income", "High income: nonOECD", 
                                            "High income: OECD")

CS_data <- CS_data %>% mutate(CS_rate_100 = CS_rate*100)
```

**1. Make a scatter plot between `CS_rate_100` and `GDP_2006`:**

```{r scatterplot}
# use this code chunk to write your code.
```

In your plot, you might notice that many of the points are condensed towards the 
lower left corner. And you might recall from the lab and assignment that the 
distributions of both cesarean delivery rate and GDP covered a wide range of 
values. Both of these variables are good candidates for log transformations to
spread out the range of data at the lowest levels.

**2. Using the `mutate()` function, add two new logged variables to the data set
`CS_data`. Call the variables `log_CS` and `log_GDP`. Use base e, also known as 
natural logarithms, to create the logged variables:**

```{r create-logged variables}
# use this code chunk to write your code.
```

**3. Remake the scatter plot using the logged variables**:

```{r scatter-logged}
# use this code chunk to write your code.
```

**4. A `geom` that you have not yet learnt is `geom_smooth()`. This geom can fit 
a curve to the data. Extend you `ggplot()` code by adding geom_smooth() to it:**

```{r add-smoothed-curve}
# use this code chunk to write your code.
```

**5. Does the relationship between logged GDP and logged CS look linear?**

Replace this text with your answer. 

**6. Modify your scatter plot by linking the color of the points to the variable
`Income_Group`.**

```{r color-points-by-income-group}
# use this code chunk to write your code.
```

**Based on this colored scatter plot, it looks like the relationship is linear 
for those countries that are not categorized as one of the two high income
categories.**

**7. For this lab, we would like to use linear regression. To do this, use a 
`dplyr` function to make a new data set called `CS_data_sub` that only contains
the low-, lower-middle, and upper-middle income countries (hint: You might want
to look at the data to see exactly what these levels are called in the data set):**

```{r subset-low-mid-income}
# use this code chunk to write your code.
```

**8. Remake the last scatter plot. This time using `CS_data_sub` to see if the 
relationship looks approximately linear between the logged variables:**

```{r scatter-low-mid}
# use this code chunk to write your code.
```

**9. Given that the relationship is approximately linear, use linear regression to 
model the relationship between `log_CS` as the response variable and `log_GDP` 
as the explanatory variable. Don't forget to specify the correct data set!:**

```{r linear-model}
# use this code chunk to write your code.
```

**10. Interpret the slope estimate:** Replace this text with your answer.

**11. Estimate what the cesarean delivery rate would be for a country with a GDP 
of 2000. Outline the steps you take to calculate your answer and provide an 
interpretation. Round your final answer to one decimal place.**

Replace this text with your answer.

**12. Is it appropriate to use the model to predict the cesarean delivery rate 
for a country with a GDP of 50,000? Why or why not? Based on the relationship in 
the full data set, would you expect the linear model to over or under predict?**

Replace this text with your answer.

\newpage 

**Submission** 

Unlike previous labs and assignments, please submit your lab *directly* to 
Gradescope. You can do this by knitting your file and downloading the PDF to 
your computer. Then navigate to Gradescope.com to submit your assignment. Here 
is a tutorial if you need help: https://www.gradescope.com/get_started. 
Scroll down on that page to "For students:submitting homework".