---
title: "Lab 2"
author: "Your name and student ID"
date: "today's date"
output: pdf_document
---

### Instructions

This lab uses the same data set as Lab 1 and Assignment 1. In those exercises, you
investigated the distribution of global cesarean delivery rates and country's 
GDPs. In part I of this lab, we will look at the relationship between these 
variables using the skills we learned from Chapters 3 and 4.

Start by loading the required libraries, reading in the data and adding on a 
variable:

```{r initial-setup}
library(dplyr)
library(ggplot2)
library(readr)
library(broom)

CS_data <- read_csv("../Data/Cesarean.csv")

# This code re-orders the variable Income_Group in the specified order.
# Note that it *does not* change the order of the data frame (like arrange() does)
# Rather, it specifies the order the data will be plotted. 
# This will make more sense when we plot the data using Income_Group, and then
# again using Income_Group_order
CS_data$Income_Group <- forcats::fct_relevel(CS_data$Income_Group, 
                                            "Low income", "Lower middle income", 
                                            "Upper middle income", "High income: nonOECD", 
                                            "High income: OECD")

CS_data <- CS_data %>% mutate(CS_rate_100 = CS_rate*100)
```

**1. Make a scatter plot between `CS_rate_100` and `GDP_2006`:**

```{r scatterplot}
#solution 
ggplot(data = CS_data, aes(x = GDP_2006, y = CS_rate_100)) + 
  geom_point()
```

In your plot, you might notice that many of the points are condensed towards the 
lower left corner. And you might recall from the lab and assignment that the 
distributions of both cesarean delivery rate and GDP covered a wide range of 
values. Both of these variables are good candidates for log transformations to
spread out the range of data at the lowest levels.

**2. Using the `mutate()` function, add two new logged variables to the data set
`CS_data`. Call the variables `log_CS` and `log_GDP`. Use base e, also know as 
natural logarithms, to create the logged variables:**

```{r create-logged variables}
#solution
CS_data <- CS_data %>% mutate(log_CS = log(CS_rate_100), 
                              log_GDP = log(GDP_2006))
```

**3. Remake the scatter plot using the logged variables**:

```{r scatter-logged}
#solution
ggplot(CS_data, aes(x = log_GDP, y = log_CS)) + 
  geom_point()
```

**4. A `geom` that you have not yet learnt is `geom_smooth()`. This geom can fit 
a curve to the data. Extend you `ggplot()` code by adding geom_smooth() to it:**
```{r add-smoothed-curve}
#solution
ggplot(CS_data, aes(x = log_GDP, y = log_CS)) + 
  geom_point() +
  geom_smooth()
```

**5. Does the relationship between logged GDP and logged CS look linear?**

Solution: It looks linear for the first 75% of the data. After that the slope 
flattens out. So overall the relationship is not linear.

**6. Modify your scatter plot by linking the color of the points to the variable
`Income_Group`.**

```{r color-points-by-income-group}
#solution
ggplot(CS_data, aes(x = log_GDP, y = log_CS)) + 
  geom_point(aes(col = Income_Group)) +
  geom_smooth()
```

**Based on this colored scatter plot, it looks like the relationship is linear 
for those countries that are not categorized as one of the two high income
categories.**

**7. For this lab, we would like to use linear regression. To do this, use a 
`dplyr` function to make a new data set called `CS_data_sub` that only contains
the low-, lower-middle, and upper-middle income countries (hint: You might want
to look at the data to see exactly what these levels are called in the data set):**

```{r subset-low-mid-income}
#solution
CS_data_sub <- CS_data %>% filter(Income_Group %in% c("Low income", 
                                                      "Lower middle income",
                                                      "Upper middle income"))

#alternative solution (remember "!" is read as "not")
CS_data_sub <- CS_data %>% 
  filter(!(Income_Group %in% c("High income: nonOECD","High income: OECD"))
         )
```

**8. Remake the last scatter plot, this time using `CS_data_sub` to see if the 
relationship looks approximately linear between the logged variables:**

```{r scatter-low-mid}
# use this code chunk to write your code.

#solution
ggplot(CS_data_sub, aes(x = log_GDP, y = log_CS)) + 
  geom_point(aes(col = Income_Group)) +
  geom_smooth()
```

**9. Given that the relationship is approximately linear, use linear regression to 
model the relationship between `log_CS` as the response variable and `log_GDP` 
as the explanatory variable. Don't forget to specify the correct data set!:**

```{r linear-model}
#solution
lm(formula = log_CS ~ log_GDP, data = CS_data_sub)

#alternative solution
mod_CS_GDP <- lm(formula = log_CS ~ log_GDP, data = CS_data_sub)
tidy(mod_CS_GDP)
```

**10. Interpret the slope estimate:**
A one unit increase in the natural log of GDP is associated with a 0.82 unit 
increase in the natural log of the cesarean delivery rate.

**11. Estimate what the cesarean delivery rate would be for a country with a GDP 
of 2000. Outline the steps you take to calculate your answer and provide an 
interpretation. Round your final answer to one decimal place.**

Solution:
We first need to convert the GDP onto the log (base $e$) scale. We can use the 
`log()` function in R to take the logarithm of 2000 using base $e$. To use base 
10, you can use the log10() function.

* Step 1. Take the natural log of GDP: log(2000) = 7.600902
* Step 2. Plug the logged GDP into the line of best fit formula:
    - Predicted logged CS = Intercept + slope*(logged GDP)
    - Predicted logged CS = -3.9405 + 0.8193*7.600902
    - Predicted logged CS = 2.286919
* Step 3. We now have the predicted logarithm of cesarean delivery. Thus, we need
to exponentiate both sides to get back to the cesarean delivery rate:
    - Predicted CS = exp(2.286919), using the property that exp(log(CS)) = CS, for natural logs.
    - Predicted CS = 9.84456% = 9.8%
* Step 4. A country with a GDP of 2000 is expected to have a cesarean delivery rate of
9.8%.

**12. Is it appropriate to use the model to predict the cesarean delivery rate 
for a country with a GDP of 50,000? Why or why not? Based on the relationship in 
the full data set, would you expect the linear model to over or under predict?**

Solution:
A GDP of 50,000 corresponds to a logged GDP of 10.82. This value is outside of
the range of our data that was used for the linear model. Thus, we should not 
use the model to make this prediction because it would be extrapolation. If we 
did use the model it would likely over predict the country's cesarean delivery 
rate because the relationship between GDP and cesarean delivery rate is less
steep for higher income countries.

\newpage 

# Part II: Simulation as tool for understanding statistics

This section contains slightly advanced material for those interested in 
learning about using R for statistical simulations.

One very useful feature of R is the ability to create simulations, both as
research and educational tools. More specifically, one can define a set of 
functions to generate "realistic" data, and then observe the behavior of 
algorithms used to estimate statistics based on the simulated data. The benefit 
of simulating data vs. using actual data, is that you now know the true values 
the statistics should take because you defined them yourself. (The downside of 
simulations, is they usually are an over-simplistic representation of the real 
process that generated the data). Simulations are used for a wide variety of
reasons. For example, you could use a simulation to investigate whether the spread
of an infectious disease that has occurred in the real world is aligned with a
model that you have proposed. 

This part of the lab begins to introduce you to the steps followed to simulate
data using R. Later in the course, we will use simulations to discuss estimation, 
inference and sampling distributions. 

For now, we generate data from a linear model $Y = a + bX + e$, where $e$ is a 
random error (otherwise, all points would lie exactly on the line).

```{r define-functions}
# runif() makes n copies of random number between 0 and 2
# Run this line and then take a look at X by printing it to the screen
X <- runif(n = 100, min = 0, max = 2)
X # Notice that there are 100 values between 0 and 2. 

# Make a random error that is normally distributed 
# We will learn more about the normal distribution in Chapter 11.
e <- rnorm(n = 100, mean = 0, sd = 0.25)

# Make Y, the reponse variable, a function of X and e
sim_intercept <- 0 # we arbitrarily set the intercept to 0 to begin. 
sim_slope <- 0.5 # we arbitrarily set the slope to 0.5 to begin. 
Y <- sim_intercept + sim_slope*X + e
dat_sim <- data.frame(Y, X)
```

**11. Make a scatter plot of Y versus X, variables in the data set `dat_sim`.**
```{r plot-the-data}
#solution
ggplot(data = dat_sim, aes(x = X, y = Y)) + 
  geom_point()
```

**12. Run a linear regression of Y on X.**
```{r run-regression}
# solution
lm(formula = Y ~ X, data = dat_sim)
```

**13. Investigate what happens if you specify `sim_slope` to equal -1 rather than
0.5. You can do this by overwriting the contents of `sim_slope` so that it equals 
-1 and then re-creating `Y` and `dat_sim` by copying and pasting the lines of 
code that define `Y` and `dat_sim`, making sure to run the new lines of code.
Also re_create the plot and re-run the linear regression. What changes?**

```{r rinse-and-repeat}
#solution
sim_intercept <- 0 #note: you don't have to re-run this specific line, since the intercept stays the same
sim_slope <- -1
Y <- sim_intercept + sim_slope*X + e
dat_sim <- data.frame(Y,X)

ggplot(data = dat_sim, aes(x = X, y = Y)) + 
  geom_point()

lm(formula = Y ~ X, data = dat_sim)
```

Solution: The pattern changes from a positive to negative trend.  Linear regression on the 
data also goes from a positive slope estimate for $X$ to a negative one.

