---
title: "Hit the dot: Dominant vs. non-dominant hand. Is there a difference?"
output: pdf_document
---

Go to this [website](https://faculty.washington.edu/chudler/java/rldot.html)

Overview of the game: See how many dots you can hit in the grid within 30 seconds. We will
each try this once with our dominant hand and once with our non-dominant hand 
(**where your dominant hand is the one you prefer to operate a computer mouse or
track pad with**).

Instructions:
1. Flip a coin to see which hand to play the game with first:
    - Heads = dominant hand first
    - Tails = non-dominant hand first

  Alternatively, rather than flipping a coin you can use `r rbinom(1, 1, 0.5)` to simulate a coin flip.

2. Push the **Start Game** button. It will start a timer counting down from 30 seconds.
During that time use only the specified hand to click the moving dot as fast as 
you can. After 30 seconds, the game will stop and display the number of dots that 
you hit. Record that number in the shared google sheet for your lab (links shown 
below). **Make sure you put it in the correct column!**. Also fill out the last column of the dataset "Dominant_hand_first". Set this variable to TRUE if you used your dominant hand in the first game or FALSE if you used your non-dominant hand in the first game.

# Lab 101B: https://docs.google.com/spreadsheets/d/1V9HCFIiXu5OtYV15DgHzBh1K6c0Rey-cFtmbmQnGiTQ/edit?usp=sharing
# Lab 102B: https://docs.google.com/spreadsheets/d/1YdBK_IMSVBiF2mKY37QvlDF60usomwD0BIxLueTZxW0/edit?usp=sharing
# Lab 103B: https://docs.google.com/spreadsheets/d/1aXMAnEzbmIRk6CG8i3GiUbt2I8Qr6Hib8rugfHuz5Qk/edit?usp=sharing
# Lab 104B: https://docs.google.com/spreadsheets/d/1i3-rxNqeK__dLHlF5rXJd6kPkFGgt4djV9pd43oTnpU/edit?usp=sharing
# Lab 105B: https://docs.google.com/spreadsheets/d/1j4SEju5-Zw_FSVFhvg1tysV-LpAE4PsQ4a_JgpZ9pK4/edit?usp=sharing
# Lab 106B: https://docs.google.com/spreadsheets/d/12BzECQ1sOT2UYvU_rKMqgsFCMFBpV5gaxkf3YWDEH9s/edit?usp=sharing
# Lab 107B: https://docs.google.com/spreadsheets/d/1d1zVuFwtpsKPthuB6XWYHrJgM2v4Z4Ok7xJYzophESU/edit?usp=sharing
# Lab 108B: https://docs.google.com/spreadsheets/d/1FtjtTpCJ7IdoMwXYHQ5GGgFw_0-VSpcbwPbAx937JuA/edit?usp=sharing
# Lab 109B: https://docs.google.com/spreadsheets/d/1uToXGdBNlhh1ltXzO-VIvyDTIAPL_GKLKznLTRvgykc/edit?usp=sharing
# Lab 110B: https://docs.google.com/spreadsheets/d/1v9Mvm2hAOB3orINrcbVldJaH1UHcsotynn91Ppyi7OY/edit?usp=sharing

3. Re-do the game, this time with the other hand. Record the results in the 
spreadsheet. 

4. Read the data from the google sheet into R. Uncomment the lines of code below.
Note that in the third line of code below you need to update the `x` argument
to be your specific lab in order to pull the correct dataset.

```{r read-into-r}
#library(googlesheets)
#library(dplyr)
#shared_gs <- gs_title(x = "Lab9_1XXB") #PLEASE REPLACE XX ACCORDING TO YOUR OWN SECTION 
# a browser will open asking you to authenticate who you are. 
# AUTHENTICATE USING YOUR BERKELEY ACCOUNT, OR YOU WON'T HAVE ACCESS!
# You can change the sheet title to you lab if you'd like. 
#our_sheet <- shared_gs %>% gs_read("Sheet1")
```

5. These data are very naturally paired. What two assumptions do we need
to make to use a paired t-test? For each assumption, either write why you think 
the assumption is met (or not met), or investigate the assumption by creating a plot,
and comment on whether the plot supports the assumption.

Answer:

```{r investigate-assumption}
# Write your code here to investigate the other assumption. Hint: You need to first compute a new variable using 
# dplyr before you make your plot :) . Don't forget to load your libraries!

```


6. Before performing the test, take a look at the data by making a "dumbbell" 
plot. This type of plot has student name on the y-axis, and the number of dots 
hit on the x axis. For each student you put a point at the two reaction times 
and connect them with a line. Here is the code to make the plot. We can also 
color the points by hand dominance. Based on the plot, comment on whether there 
appears to be a significant difference between the number of points hit between
the dominant and non-dominant hand.

Here is the code to make the dumbbell chart. You will need to change `dat` to 
the name of your saved dataset.

```{r make-dumbell-chart}
# You are not expected to know how to make this plot yourself!

# ggplot(data = dat, aes(x = Dominant_num_dots_hit, y = Student_name)) +
#   geom_segment(aes(xend = Non_dominant_num_dots_hit, yend = Student_name)) + 
#   geom_point(aes(col = "Dominant")) +
#   geom_point(aes(x = Non_dominant_num_dots_hit, col = "Non-dominant"))
```

Comment:

7. Use R to conduct a paired two-sided t-test on the data, and the 95% confidence
interval for the test. Interpret the p-value and the confidence interval for the 
test. Hint: see the code from the paired t-test (Monday's lecture on the rest of 
Chapter 17). You can use one function to both calculate the p-value for the 
two-sided test and create a 95% confidence interval for the difference:

```{r conduct-paired-t-test}
#Your code here
```

Answer:



8. Re-run the code for the test, but this time set `paired=F`, which is incorrect.
The reason we want to run the incorrect test is to compare the p-value from this test
to the p-value from the paired t-test. Is it smaller or larger? Why is that?

```{r}
#Your code here
```

Answer: 

9. Lastly, we didn't use the data on the last column in the data frame, which recorded
whether you were randomized to using your dominant hand first. Why might this matter?
What could we have done to investigate whether it mattered?

Answer:


**Submission** 

Please submit this lab file *directly* to Gradescope under the "Lab 9" 
Assignment. You can do this by 
knitting your file and downloading the PDF to your computer. Then navigate to 
Gradescope.com to submit your assignment. Here is a tutorial if you need help: https://www.gradescope.com/get_started. Scroll down on that page to 
"For students:submitting homework".
