---
title: "Hit the dot: Dominant vs. non-dominant hand. Is there a difference?"
output: pdf_document
---

Go to this [website](https://faculty.washington.edu/chudler/java/rldot.html)

The game: See how many dots you can hit in the grid within 30 seconds. We will
each try this once with our dominant hand and once with our non-dominant hand 
(**where your dominant hand is the one you prefer to operate a computer mouse or
track pad with**).

Instructions:
1. Flip a coin to see which hand to play the game with first:
    - Heads = dominant hand first
    - Tails = non-dominant hand first

  Rather than flipping a coin you can use `r rbinom(1, 1, 0.5)` to simulate a coin flip!

    
2. Push the **Start Game** button. It will start a timer counting down from 30 seconds.
During that time use only the specified hand to click the moving dot as fast as 
you can. After 30 seconds, the game will stop and display the number of dots that 
you hit. Record that number in the shared google sheet. **Make sure you put it 
in the correct column!**. Also fill out the last column of the dataset 
"Dominant_hand_first". Set this variable to TRUE if you used your dominant hand 
in the first game or FALSE if you used your non-dominant hand in the first game.

3. Re-do the game, this time with the other hand. Record the results in the 
spreadsheet. 

4. Read the data from the google sheet into R. 

# Lab 101B: https://docs.google.com/spreadsheets/d/1V9HCFIiXu5OtYV15DgHzBh1K6c0Rey-cFtmbmQnGiTQ/edit?usp=sharing
# Lab 102B: https://docs.google.com/spreadsheets/d/1YdBK_IMSVBiF2mKY37QvlDF60usomwD0BIxLueTZxW0/edit?usp=sharing
# Lab 103B: https://docs.google.com/spreadsheets/d/1aXMAnEzbmIRk6CG8i3GiUbt2I8Qr6Hib8rugfHuz5Qk/edit?usp=sharing
# Lab 104B: https://docs.google.com/spreadsheets/d/1i3-rxNqeK__dLHlF5rXJd6kPkFGgt4djV9pd43oTnpU/edit?usp=sharing
# Lab 105B: https://docs.google.com/spreadsheets/d/1j4SEju5-Zw_FSVFhvg1tysV-LpAE4PsQ4a_JgpZ9pK4/edit?usp=sharing
# Lab 106B: https://docs.google.com/spreadsheets/d/12BzECQ1sOT2UYvU_rKMqgsFCMFBpV5gaxkf3YWDEH9s/edit?usp=sharing
# Lab 107B: https://docs.google.com/spreadsheets/d/1d1zVuFwtpsKPthuB6XWYHrJgM2v4Z4Ok7xJYzophESU/edit?usp=sharing
# Lab 108B: https://docs.google.com/spreadsheets/d/1FtjtTpCJ7IdoMwXYHQ5GGgFw_0-VSpcbwPbAx937JuA/edit?usp=sharing
# Lab 109B: https://docs.google.com/spreadsheets/d/1uToXGdBNlhh1ltXzO-VIvyDTIAPL_GKLKznLTRvgykc/edit?usp=sharing
# Lab 110B: https://docs.google.com/spreadsheets/d/1v9Mvm2hAOB3orINrcbVldJaH1UHcsotynn91Ppyi7OY/edit?usp=sharing

# sample: https://docs.google.com/spreadsheets/d/1v9Mvm2hAOB3orINrcbVldJaH1UHcsotynn91Ppyi7OY/edit?usp=sharing

```{r read-into-r}
library(googlesheets)
library(dplyr)
shared_gs <- gs_title(x = "Lab9_sampleB") 
# a browser will open asking you to authenticate who you are. 
# AUTHENTICATE USING YOUR BERKELEY ACCOUNT, OR YOU WON'T HAVE ACCESS!
# You can change the sheet title to you lab if you'd like. 
our_sheet <- shared_gs %>% gs_read("Sheet1")
```

5. These data are very naturally paired. What two assumptions do we need
to make to use a paired t-test? For each assumption, either write why you think 
the assumption is met (or not met), or investigate the assumption by creating a plot,
and comment on whether the plot supports the assumption.

Solution:
- Paired differences in the population are Normally distributed. Can investigate
this by looking at the distribution of the paired differences using a histogram
or QQ plot.
- Sample is a SRS from the population. Who is the population here? We didn't specify!
If the population is the ~300 PH142 students can you consider your lab section
as a SRS? Why or why not? What about if the population is all students at 
Berkeley? Even if not a SRS, can we treat it like one given the outcome we're 
interested in?

```{r investigate-assumption}
# Write your code here to investigate the other assumption. Hint: You need to first compute a new variable using 
# dplyr before you make your plot :) . Don't forget to load your libraries!
# Solution 
library(ggplot2)
our_sheet <- our_sheet %>% mutate(difference = Dominant_num_dots_hit - Non_dominant_num_dots_hit)
ggplot(data = our_sheet, aes(x = difference)) + geom_histogram(binwidth = 2, col = "white") #may need to chose a better `binwidth`
```

Interpret the histogram: With so little data it is hard to assess Normality, but we can say that there are no really distant outliers based on this sample and using a binwidth of 2 these data look "Normal-ish".

6. Before performing the test, take a look at the data by making a "dumbbell" 
plot. This type of plot has student name on the y-axis, and the number of dots 
hit on the x axis. For each student you put a point at the two reaction times 
and connect them with a line. Here is the code to make the plot. We can also 
color the points by hand dominance. Based on the plot, comment on whether there 
appears to be a significant difference between the number of points hit between
the dominant and non-dominant hand.

Here is the code to make the dumbbell chart. You will need to change `dat` to 
the name of your saved dataset.

```{r make-dumbell-chart}
# This code is provided to students because it is a bit advanced.
# You are not expected to know how to make this plot yourself!
ggplot(data = our_sheet, aes(x = Dominant_num_dots_hit, y = Student_name)) +
  geom_segment(aes(xend = Non_dominant_num_dots_hit, yend = Student_name)) + 
  geom_point(aes(col = "Dominant")) +
  geom_point(aes(x = Non_dominant_num_dots_hit, col = "Non-dominant"))
```

Solution: For these data, you can see that all students except for 1 had a higher number of dots hit with their dominant vs. non-dominant hand. That is for 18 out of 19 students, there is a difference in the number of dots hit, so we would expect that the tests would reject a null hypothesis of no difference.

7. Use R to conduct a paired two-sided t-test on the data, and the 95% confidence
interval for the test. Interpret the p-value and the confidence interval for the 
test. 

```{r conduct-paired-t-test}
# Solution 
t.test(our_sheet %>% pull(Dominant_num_dots_hit), 
       our_sheet %>% pull(Non_dominant_num_dots_hit),
       paired = T, alternative = "two.sided")
```

Solution:

The p-value for the test is <0.0001. This means that under the null hypothesis of no difference between the hands in the number of dots hit, there is <0.0001% chance of observing the difference that we saw (or one of greater magnitude) between the sample average of the individual differences. This probability is so small that we reject the null hypothesis in favour of the alternative hypothesis that there is a difference between the hands in the number of dots hit.

The 95% CI for the difference is 4.05 to 7.63. If our model assumptions are correct and there is
only random error affecting the estimate, this method for calculating confidence intervals will contain
the true underlying value of difference between dominant and non-dominant hands 95% of the time (19 times out of 20).

8. Re-run the code for the test, but this time set `paired=F`, which is incorrect.
The reason we want to run the incorrect test is to compare the p-value from this test
to the p-value from the paired t-test. Is it smaller or larger? Why is that?

```{r}
# Solution
t.test(our_sheet %>% pull(Dominant_num_dots_hit), 
       our_sheet %>% pull(Non_dominant_num_dots_hit),
       paired = F, alternative = "two.sided")
```

Solution A (based on specific data): Using these data, the p-value is larger. Here the p-value was still a very small number, and still provides evidence against the null hypothesis of no difference between the groups. This is because the mean number of dots hit for the dominant hand (28.4) was so much larger than for the non-dominant hand (22.6) that the test still detected the difference, even though we forgot to tell the test that these data were paired. 

Solution B (generic): The p-value for the independent test is larger. This means there is less evidence (or maybe even minimal/"no") evidence against the null hypothesis of no difference. It is larger/there is less evidence because the independent test does not account for the matching in the data. So it included all the "noise" due to differences between individuals. 

9. Lastly, we didn't use the data on the last column in the data frame, which recorded
whether you were randomized to using your dominant hand first. Why might this matter?
What could we have done to investigate whether it mattered?

Solution B: It could be that there was a "learning effect" and that when you played the game with the first hand, you partially struggled because it was your first time playing the game. If that was the case, then you would expect to do a bit better using your second hand because you are more familar with the game. To investigate this, we could have made a histogram of the individual differences for those who used their dominant hand first, and compared it to a separate histogram for those who used their non-dominant hand first. If these plots look very different this might suggest that there was a learning effect. 

Solution A: It could be that playing the game had a "learning" effect, meaning that you learnt how to play it with the first hand, which means that you become better at playing it with the second hand. If this was the case, then those that use their dominant hand second are expected to perform better with that hand than those who used their dominant hand first. To investigate this. We could make a boxplot for the number of dots hit with the first hand, and a second box plot for the second hand, and do this *separately* according to which hand was first. We will see an example of this in lecture.


**Submission** 

Please submit this lab file *directly* to Gradescope under the "Lab 9" 
Assignment. You can do this by 
knitting your file and downloading the PDF to your computer. Then navigate to 
Gradescope.com to submit your assignment. Here is a tutorial if you need help: https://www.gradescope.com/get_started. Scroll down on that page to 
"For students:submitting homework".
