---
title: 'Assignment 1 Solutions'
output: pdf_document
---

### Sleep times in mammals

The file `sleep.csv` contains the sleeptimes and weights for a set of mammals. 
Use the following code to read the file into memory and store it in the R object
called sleep:

```{r import-and-save-data}
library(tidyverse)
sleep <- read_csv("../../Data/sleep.csv")
```

Notice that an object called `sleep` appeared in the Environment tab. Use four
useful functions discussed in lecture to examine the sleep data set:

```{r four-functions}
# solution

dim(sleep)
names(sleep)
head(sleep)
str(sleep)
```

| Column name	| Description |
|-------------|-------------|
|name	|common name|
|genus	|taxonomic rank|
|vore	|carnivore, omnivore or herbivore?|
|order	| taxonomic rank|
|conservation	|the conservation status of the mammal|
|sleep_total|	total amount of sleep, in hours|
|sleep_rem |	Rapid eye movement (REM) sleep, in hours |
|sleep_cycle|	length of sleep cycle, in hours|
|awake|	amount of time spent awake, in hours|
|brainwt|	brain weight in kilograms|
|bodywt|	body weight in kilograms|

1) Select a set of columns. Specifically select the `awake`, `brainwt`, and `bodywt` columns. Assign this smaller dataset to a data frame called `smaller_sleep_data`

```{r}
# solution
smaller_sleep_data <- sleep %>% select(awake, brainwt, bodywt)

#alternative solution:
smaller_sleep_data <- select(sleep, awake, brainwt, bodywt)
```

2) To select a range of columns by name, use the “:” (colon) operator. Redo the 
selection for question 1, but use the colon operator:

```{r}
# solution
#note that this only works because these variables appear side-by-side in the dataset
smaller_sleep_data <- sleep %>% select(awake:bodywt)
```

3) Select all the columns except for the `vore` variable

```{r}
# solution
sleep %>% select(-vore)
```

4) Run the following chunk of code. What does it return?

```{r}
select(sleep, starts_with("sl"))
```

Solution: It returns the columns that start with sl: `sleep_total`, `sleep_rem`, `sleep_cycle`

5) Rewrite the previous chunk of code using the pipe operator.

```{r}
# solution

sleep %>% select(starts_with("sl"))
```

6) Filter the rows for mammals that sleep a total of more than 16 hours.

```{r}
#solution

sleep %>% filter(sleep_total > 16)
```

7) Filter the rows for mammals that sleep a total of more than 16 hours and have
a body weight of greater than 1 kilogram.

```{r}
#solution

sleep %>% filter(sleep_total > 16 & bodywt > 1)

# alternative solution

sleep %>% filter(sleep_total > 16, bodywt > 1)
```


8) Suppose you are specifically interested in the sleep of horses and giraffes.
Write a line of code to print the data frame for horses and giraffes only:

```{r}
#solution
sleep %>% filter(name %in% c("Horse", "Giraffe")) #Notice: quotes and capitalization!
```

9) Order the dataset by sleep time from shortest sleep time to longest sleep time

```{r}
#solution
sleep %>% arrange(sleep_total)
```

10) Now order for longest sleep time to shortest sleep time

```{r}
#solution
sleep %>% arrange(-sleep_total)
```

11) Suppose you are interested in the order of sleep time, but according to whether
the animal is a carnivore, herbivore, insectivore, or omnivore. Rewrite the above statement
to order sleep time according to the type of "-vore" that then animal is:

```{r}
#solution
sleep %>% arrange(vore, -sleep_total)
```

12) Create a new column called `rem_proportion` which is the ratio of rem sleep to total amount of sleep.

```{r}
#solution
sleep %>% mutate(rem_proportion = sleep_rem/sleep_total)
```

13) Copy your previous line of code. Revise it to add a second column called bodywt_grams which is the bodywt column in grams.

```{r}
#solution
sleep %>% mutate(rem_proportion = sleep_rem/sleep_total, bodywt_grams = bodywt * 1000)
```

14) Calculate the average sleep time across all the animals in the dataset using
a dplyr function. Call the new variable `sleep average`:

```{r}
#solution
sleep %>% summarize(sleep_average = mean(sleep_total))
```

15) Calculate the average sleep time for each type of "-vore". Hint: you'll need
to use two dplyr functions!

```{r}
#solution
sleep %>% 
  group_by(vore) %>%
  summarize(sleep_average = mean(sleep_total))
```


